---
title: "CRI-BIO-XXX: Demostration"
author: "CRI Bioinformatics Core"
date: "_`r Sys.Date()`_"
output:
  html_document:
    code_folding: show
    df_print: paged
    theme: yeti
    highlight: tango
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
    number_sections: yes
    self_contained: true
params:
  raw_data_root_folder: "raw_data"
  sampleinfo: "sampleinfo.csv" # The first column must be Sample_Name whose values are exactly the same as those in metadata files (in .csv format) of the raw data; metadata files should be inside the raw_data_root_folder; the samples in sampleinfo can be a subset of raw data. There must be a Group column indicating sample groups to compare. 
  platform: "EPICv2"
  result_folder: "Demostration"
  remove_bad_samples: TRUE
  unmask: FALSE
  tSNE_perplexity: 3
  skip_DM: FALSE
  test_mode: TRUE
  contrast_pairs: !expr list(
    c("STK11 Adnexal Tumor", "Adult Granulosa Cell Tumor"),
    c("STK11 Adnexal Tumor", "Mesothelioma")
    )
  species: "human"
  ora_fdr_thres: 0.05
---

```{r setup, include=FALSE}

library(ggplot2)
library(plotly)
library(patchwork)
library(Rtsne)
library(matrixStats)
library(dplyr)
library(tibble)
library(stats)
library(circlize)
library(dendextend)
library(VIM)
library(sesame)
library(ComplexHeatmap)
library(ggrepel)
library(org.Hs.eg.db)
library(org.Mm.eg.db)
library(clusterProfiler)

sesameDataCache()

knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

if(!dir.exists(params$result_folder)) dir.create(params$result_folder)

```

# Experiment design

We use [sesame](https://bioconductor.org/packages/release/bioc/vignettes/sesame/inst/doc/sesame.html) to analyze the data as recommended by Illumina.

We use the standard openSesame workflow to process raw signal data to beta values. A description of the method details can be found in [this article](https://epicom.biomedcentral.com/articles/10.1186/s43682-023-00021-5). 

```{r pre-processing}

if (file.exists(paste0(params$result_folder, ".RData"))){
  cat(paste0(params$result_folder, ".RData"), " already exists in the working directory. Use it for analyses downstream of QC instead of reading from raw data. If you want to restart from reading in raw data, please remove ", paste0(params$result_folder, ".RData"), " from current working directory and knit this script again. ")
  load(paste0(params$result_folder, ".RData"))
}else{
  metadata = list.files(params$raw_data_root_folder, pattern = ".csv", full.names = T, recursive = T)
  metadata = lapply(metadata, function(x) read.csv(x, skip = 5))
  metadata = do.call(rbind, metadata)

  sampleinfo = read.csv(params$sampleinfo, na.strings = "")

  coldata = merge(metadata, sampleinfo, by = "Sample_Name")
  coldata$IDAT = paste(coldata$Sentrix_ID, coldata$Sentrix_Position, sep = "_")
  coldata = coldata[,c("IDAT", "Sample_Name", names(sampleinfo)[-1])]

  if(nrow(metadata)>nrow(coldata))  cat("There are ", nrow(metadata), " samples found in the raw data folder, but only ", nrow(coldata), " samples have experimental condition records. We will omit samples without any experimental condition record from the downstream analysis. ")

  idats = searchIDATprefixes(params$raw_data_root_folder)
  
  sdfs = lapply(idats, readIDATpair, platform = params$platform) %>% setNames(gsub(".*/", "", idats))
  sdfs = sdfs[coldata$IDAT]

  betas = openSesame(sdfs, collapseToPfx=FALSE)
  betas_unmask = openSesame(sdfs, collapseToPfx=FALSE, mask=FALSE)
  coldata = coldata[na.omit(match(colnames(betas), coldata$IDAT)),]
}

```

# Experiment design

## Sample information for each IDAT pair

```{r table1}
if (file.exists(paste0(params$result_folder, ".RData"))){
  sample_dt
}else{
  sample_dt = DT::datatable(coldata, rownames = FALSE)
  sample_dt
}

```

## Sample counts per condition

```{r table2}
if (file.exists(paste0(params$result_folder, ".RData"))){
  freq_dt
}else{
  freq_dt = coldata[, -c(1, 2), drop = F] %>% group_by(across(everything())) %>% summarise(Frequency = n()) %>% ungroup() %>%  as.data.frame() %>% DT::datatable(rownames = FALSE)
  freq_dt
}
```

# QC

## Tables

```{r QC}

if (!file.exists(paste0(params$result_folder, ".RData"))){
  qcs = list()
  qcs_df = list()
  qcs_dt = list()
  # rankstats = list()
  # rankstats_df = list()
  # rankstats_dt = list()
  for (f in c("detection", "intensity", "numProbes", "channel", "dyeBias", "betas")) {
    qcs[[f]] = openSesame(sdfs, prep="", func=sesameQC_calcStats, funs=f)
    qcs_df[[f]] = do.call(rbind, lapply(qcs[[f]], as.data.frame))
    colnames(qcs_df[[f]]) = qcs[[f]][[1]]@group[[1]][colnames(qcs_df[[f]])]
    qcs_dt[[f]] = DT::datatable(data = qcs_df[[f]], caption = names(qcs[[f]][[1]]@group), rownames = T)
    # rankstats[[f]] = lapply(qcs[[f]], sesameQC_rankStats, platform=params$platform_plt)
    # rankstats_df[[f]] = do.call(rbind, lapply(rankstats[[f]], as.data.frame))
    # colnames(rankstats_df[[f]]) = rankstats[[f]][[1]]@group[[1]][colnames(rankstats_df[[f]])]
    # rankstats_dt[[f]] = DT::datatable(data = rankstats_df[[f]], caption = names(rankstats[[f]][[1]]@group), rownames = T)
  }
}

htmltools::tagList(setNames(qcs_dt, NULL))

```

## Selected barplots

Please find the full set of barplots in the `QC-Bar` folder.

```{r barplots }

if (!file.exists(paste0(params$result_folder, ".RData"))){
  if(!dir.exists(paste0(params$result_folder, "/QC-Bar"))) dir.create(paste0(params$result_folder, "/QC-Bar"))

  for (f in c("detection", "intensity", "numProbes", "channel", "dyeBias", "betas")) {
    h = length(qcs[[f]][[1]]@group[[1]])*250
    png(paste0(params$result_folder, "/QC-Bar/barplot_", f, ".png"), width = 500, height = h, units = "px")
    sesameQC_plotBar(qcs[[f]], keys = "") %>% print()
    dev.off()
  }
}

for (f in c("detection", "intensity")) {
  sesameQC_plotBar(qcs[[f]]) %>% print()
}

```

```{r RedGrnQQ, include=FALSE }

if (!file.exists(paste0(params$result_folder, ".RData"))){
  if(!dir.exists(paste0(params$result_folder, "/QC-RedGrnQQ"))) dir.create(paste0(params$result_folder, "/QC-RedGrnQQ"))

  for (f in names(sdfs)) {
    png(paste0(params$result_folder, "/QC-RedGrnQQ/qqplot_", f, ".png"), width = 500, height = 500, units = "px")
    sesameQC_plotRedGrnQQ(sdfs[[f]]) %>% print()
    dev.off()
  }
}

```

```{r IntensVsBetas, include=FALSE }

if (!file.exists(paste0(params$result_folder, ".RData"))){
  if(!dir.exists(paste0(params$result_folder, "/QC-IntensVsBetas"))) dir.create(paste0(params$result_folder, "/QC-IntensVsBetas"))

  for (f in names(sdfs)) {
    png(paste0(params$result_folder, "/QC-IntensVsBetas/intbetaplot_", f, ".png"), width = 500, height = 500, units = "px")
    sesameQC_plotIntensVsBetas(sdfs[[f]]) %>% print()
    dev.off()
  }
}

```

## Sample counts per condition after removing bad samples

```{r table3, results='asis'}

if(params$remove_bad_samples){
  samples_rm <- rownames(qcs_df$detection)[qcs_df$detection[grep("% Detection Success\\s+$",names(qcs_df$detection))]<0.4]
  
  if (length(samples_rm)>0) {
    coldata <- coldata[!coldata$IDAT %in% samples_rm,]
    betas <- betas[,!colnames(betas) %in% samples_rm]
    betas_unmask <- betas_unmask[,!colnames(betas_unmask) %in% samples_rm]
    cat("Bad samples (detection success rate < 40%) to remove:", paste(samples_rm, collapse = ", "))
    
    coldata[, -c(1, 2), drop = F] %>% group_by(across(everything())) %>% summarise(Frequency = n()) %>% ungroup() %>%  as.data.frame() %>% DT::datatable(rownames = FALSE)
  }else{
    cat("There is no bad sample. All samples have a detection success rate >= 40%.")
  }
}else{
  samples_rm <- rownames(qcs_df$detection)[qcs_df$detection[grep("% Detection Success\\s+$",names(qcs_df$detection))]<0.4]
  if (length(samples_rm)>0) {
    cat("Potential bad samples (detection success rate < 40%) to remove:", paste(samples_rm, collapse = ", "), "<br>")
    cat("For analyses in this report, we use all samples. ")
  }else{
    cat("There is no bad sample. All samples have a detection success rate >= 40%.")
  }
}

```

```{r save env}
if (!file.exists(paste0(params$result_folder, ".RData"))) save(list = ls()[-grep("params", ls())], file = paste0(params$result_folder, ".RData"))
```

# t-SNE plots

For the t-SNE analysis below, we applied perplexity = `r params$tSNE_perplexity`. 

```{r t-SNE notes, results='asis'}
if(params$unmask){
  cat("Note that we used a probe matrix without masking bad detection p-values for the t-SNE analysis. ", nrow(betas_unmask), " CpG loci were used here.")
}else{
  cat("Note that probes with bad detection p-values were masked as NAs in the QC steps. These probes were removed in the t-SNE analysis since they were unacceptable as inputs for t-SNE. ", nrow(na.omit(betas)), " out of ", nrow(betas), " CpG loci were removed here.")
}
```

## Using all CpG loci (probes)

```{r t-SNE function, include=FALSE}

tsne_plot <- function(tsne_result, coldata, title){
  tsne_data <- data.frame(
    tSNE1 = tsne_result$Y[,1],
    tSNE2 = tsne_result$Y[,2]
  )
  plot_data <- cbind(tsne_data, coldata[,-1])
  if (ncol(plot_data)==4) {
    gp <- ggplot(plot_data, aes(x = tSNE1, y = tSNE2, label = Sample_Name, color = !!sym(names(plot_data)[4]))) +
      geom_point(size = 2) +
      labs(title = title) +
      theme_minimal() +
      scale_color_brewer(palette = "Set2")
  }else if (ncol(plot_data)>4) {
    gp <- ggplot(plot_data, aes(x = tSNE1, y = tSNE2, label = Sample_Name, color = !!sym(names(plot_data)[4]), shape = !!sym(names(plot_data)[5]))) +
      geom_point(size = 2) +
      labs(title = title) +
      theme_minimal() +
      scale_color_brewer(palette = "Set2")
  }else{
    gp <- ggplot(plot_data, aes(x = tSNE1, y = tSNE2, label = Sample_Name)) +
      geom_point(size = 2) +
      labs(title = title) +
      theme_minimal() +
      scale_color_brewer(palette = "Set2")
  }
  return(gp)
}

```

```{r t-SNE all, include=FALSE}

if(params$unmask){
  betas_clean <- betas_unmask
}else{
  betas_clean <- na.omit(betas)
}
set.seed(123)  # Set a seed for reproducibility
tsne_result <- Rtsne(t(betas_clean), dims = 2, perplexity = params$tSNE_perplexity, verbose = TRUE)
gp <- tsne_plot(tsne_result, coldata, "t-SNE clustering")

```

```{r plot out all, fig.width=8, results='asis'}
ggplotly(gp, tooltip = "label")
```

## Using the top 10k most variable CpG loci (probes)

```{r t-SNE top10k, include=FALSE}

top10k <- rownames(betas_clean)[order(rowVars(betas_clean), decreasing=TRUE)] %>% head(10000)
betas_clean_top <- betas_clean[top10k,]

set.seed(123)  # Set a seed for reproducibility
tsne_result <- Rtsne(t(betas_clean_top), dims = 2, perplexity = params$tSNE_perplexity, verbose = TRUE)

gp <- tsne_plot(tsne_result, coldata, "t-SNE clustering")

```

```{r plot out top10k, fig.width=8, results='asis'}
ggplotly(gp, tooltip = "label")
```

# Heatmap with unsupervised hierarchical clustering

```{r heatmap notes, results='asis'}
if(params$unmask){
  cat("Note that we used a probe matrix without masking bad detection p-values for the heatmap and unsupervised hierarchical clustering (Ward’s method). The unsupervised hierarchical clustering was done using the top 10k probes with the highest Median Absolute Deviation (MAD) whose beta values were shown on the heatmap. Probes were also clustered with the Ward's method. ")
}else{
  cat("The unsupervised hierarchical clustering (Ward’s method) was done using the top 10k probes with the highest Median Absolute Deviation (MAD) whose beta values were shown on the heatmap. Some beta values were masked due to low confidence in the detection of signals. It’s usually not recommended to use these low-confidence values in the downstream DM analysis. For heatmap visualization purpose, we removed any probes containing an NA value. Probes were then clustered with the Ward's method. ")
}
```

```{r heatmap save, include=FALSE }

if(params$unmask){
  mad_values <- apply(betas_unmask, 1, mad)  # Calculate MAD, ignoring NAs
  betas.sel <- betas_unmask[order(mad_values, decreasing = TRUE)[1:10000], ]  # Subset the data to include only the top probes
}else{
  mad_values <- apply(betas, 1, mad, na.rm = TRUE)  # Calculate MAD, ignoring NAs
  betas.sel <- betas[order(mad_values, decreasing = TRUE)[1:10000], ]  # Subset the data to include only the top probes
  # betas.sel <- imputeBetas(betas.sel) # CANNOT do imputation since it turned out to alter the relationship between samples.
  betas.sel <- na.omit(betas.sel)
}

sample_anno <- coldata[,-c(1,2), drop = FALSE]
sample_anno[] = lapply(sample_anno, as.factor)
annotation_colors <- lapply(sample_anno, function(x) {
  colors <- c(RColorBrewer::brewer.pal(8, "Set2"), RColorBrewer::brewer.pal(9, "Set1"))[1:length(levels(x))]
  setNames(colors[seq_along(levels(x))], levels(x))
})

pdf(file.path(params$result_folder, "heatmap.pdf"), width = 8, height = 16)
hm <- Heatmap(
  betas.sel,
  name = "Beta Values",  
  col = colorRamp2(c(0, 0.5, 1), c("blue", "whitesmoke", "red")),  
  na_col = "black",  
  show_row_names = FALSE,
  cluster_rows = TRUE,
  clustering_method_columns = "ward.D2",  # Specify clustering method for columns
  clustering_method_rows = "ward.D2",
  use_raster = TRUE,
  top_annotation = HeatmapAnnotation(
    df = sample_anno,
    col = annotation_colors,  # Assign colors automatically generated for each factor
    annotation_name_side = "left"  # Position of the annotation name
  )
)
draw(hm)
dev.off()

```

```{r heatmap plot, results='asis'}
cat(paste0('<iframe src="', file.path(params$result_folder, "heatmap.pdf"), '" width="100%" height="600px"></iframe>'))
```

# Differentially methylated probes and regions

## Differentially methylated probes (DMPs)

```{r DE-test, results='asis'}

if(!params$skip_DM) {
  rownames(coldata) <- coldata$IDAT

  pval = 0.05
  eff_thres = 0.2
  seg_fdr = 0.05
  
  if(!exists("smry")) { 
    smry = list()
    rerun_DML = TRUE
  }else{
    rerun_DML = FALSE
  }
  test_result = list()
  test_result_sig = list()
  test_result_dt = list()
  merged = list()
  merged_sig = list()
  merged_dt = list()
  merged_plt = list()
  warning_flag = c()
  
  for(comp in params$contrast_pairs){
    if(nrow(coldata[coldata$Group %in% comp[1],])==1 && nrow(coldata[coldata$Group %in% comp[2],])==1) {
      replicates = FALSE
      warning_flag <- c(warning_flag, paste(comp, collapse = " vs "))
    }else{
      replicates = TRUE
    }
    
    betas.sub = betas[,coldata$Group %in% comp]
    coldata.sub = coldata[coldata$Group %in% comp,]
    
    coldata.sub$Group <- factor(coldata.sub$Group, levels = comp)
    
    betas_ok = (checkLevels(betas.sub, coldata.sub$Group))
    comp_name = paste(comp, collapse = " vs ")
    cat("<h4>", comp_name, "</h4>")
    cat("We removed probes missing a non-NA value for any factor level (", paste0(unique(coldata.sub$Group), collapse = " or "), " in this contrast). The remaining ", sum(betas_ok), " out of ", nrow(betas.sub), " probes were used for DM analysis.", sep = "")
  
    betas.sub = betas.sub[betas_ok,]
    if(params$test_mode) betas.sub = betas.sub[1:2000,] # for testing
    
    if(rerun_DML) smry[[comp_name]] = DML(betas.sub, fm = ~Group, meta = coldata.sub)
    test_result[[comp_name]] = summaryExtractTest(smry[[comp_name]])
    
    if(replicates) {
      test_result_sig[[comp_name]] = test_result[[comp_name]] %>% filter(.[[5]] < pval, .[[7]] > eff_thres) %>% arrange(desc(.[[7]]))
    }else{
      test_result_sig[[comp_name]] = test_result[[comp_name]] %>% filter(.[[7]] > eff_thres) %>% arrange(desc(.[[7]]))
    }
    
    test_result_dt[[comp_name]] = DT::datatable(data = test_result_sig[[comp_name]][1:100,], caption = paste(comp_name, "Top 100 Differential Methylated CpGs"), rownames = F)
  
    if(replicates) {
      cat("There were ", nrow(test_result_sig[[comp_name]]), " differentially methylated CpGs with a p-value < ", pval, " and an effective size > ", eff_thres, ".<br>", sep = "")
    }else{
      cat("***Note that there was no replicate for either ", comp[1], " or ", comp[2], ". So P-value was not calculated for this comparison. All the ", nrow(test_result_sig[[comp_name]]), " CpGs with an effective size > ", eff_thres, " were saved to the results tables.***<br>", sep = "")
    }
    cat("Up-regulated in", comp[2], ":", nrow(test_result_sig[[comp_name]] %>% filter(.[[3]]>0)), "<br>")
    cat("Down-regulated in", comp[2], ":", nrow(test_result_sig[[comp_name]] %>% filter(.[[3]]<0)), "<br>")
  
    if(!dir.exists(file.path(params$result_folder, "DMP/"))) dir.create(file.path(params$result_folder, "DMP/"), recursive = TRUE)
    write.csv(test_result_sig[[comp_name]], file = paste0(params$result_folder, "/DMP/DMP_", gsub("\\s", "_", comp_name), ".csv"))
    
    merged[[comp_name]] = DMR(betas.sub, smry[[comp_name]], colnames(attr(smry[[comp_name]], "model.matrix"))[2]) # merge CpGs to regions
    if(replicates) {
      merged_sig[[comp_name]] = merged[[comp_name]] %>% dplyr::filter(abs(Seg_Est) > eff_thres) %>% dplyr::filter(Seg_Pval_adj < seg_fdr) %>% arrange(desc(abs(Seg_Est)))
    }else{
      merged_sig[[comp_name]] = merged[[comp_name]] %>% dplyr::filter(abs(Seg_Est) > eff_thres) %>% arrange(desc(abs(Seg_Est)))
    }
    merged_dt[[comp_name]] = DT::datatable(data = unique(merged_sig[[comp_name]][grep("chr",merged_sig[[comp_name]]$Seg_Chrm),1:7]), caption = paste(comp_name, "Differential Methylated Regions (DMRs)"), rownames = F)
    merged_plt[[comp_name]] = unique(merged_sig[[comp_name]][grep("chr",merged_sig[[comp_name]]$Seg_Chrm),1:7])
  
    if(!dir.exists(file.path(params$result_folder, "DMR/"))) dir.create(file.path(params$result_folder, "DMR/"), recursive = TRUE)
    write.csv(merged_sig[[comp_name]], file = paste0(params$result_folder, "/DMR/DMR_", gsub("\\s", "_", comp_name), ".csv"))
  }
}

```

```{r save DML results}
if(!params$skip_DM && !params$test_mode) save(list = ls()[-grep("params", ls())], file = paste0(params$result_folder, ".RData"))
```

### Top 100 significant DMPs

The tables below show top 100 significant DMPs. Please find the full significant DMP lists in the `DMP` folder.

**Note:** With the exception of the Intercept, there are four groups of columns, each starting with "Est_", "Pval_", "FPval_", and "Eff_" as prefix. Here are what they represent:

- Est: The slope estimate (aka the β coefficient, not to be confused with the DNA methylation β-value though) for continuous variable. DNA methylation difference of the current level with respect to the reference level for nominal contrast variables. There is a special column named Est_X.Intercept. It corresponds to the base-level methylation of the reference.

- Pval: The unadjusted p-values of t-testing the slope. This represents the statistical significance of the methylation difference. The Pval_X.Intercept tests whether the reference level is significantly different from zero.

- FPval: The unadjusted p-value of the F-test contrasting the full model against a reduced model with the labeled contrast variable held out. Note that "Pval_" and "FPval_" are equivalent when the contrast variable is a 2-level factor, i.e., in the case of a pairwise comparison.

- Eff: The effect size of each normial contrast variable. This is equivalent to the maximum slope subtracted by the minimum level including the reference level (0).

```{r DMP tables}
if(!params$skip_DM) {
  htmltools::tagList(setNames(test_result_dt, NULL))
}
```

### Barplots of significant DMPs

```{r barplot-function}
DM_barplots <- function(plot_data, title){
  space = (max(plot_data$Hypermethylation)-min(plot_data$Hypomethylation))*0.15/2
  bp_ratio = nrow(plot_data)/(max(plot_data$Hypermethylation)-min(plot_data$Hypomethylation))/4
  ggplot(plot_data, aes(x = Contrast)) +
    geom_bar(aes(y = Hypermethylation, fill = "Hypermethylation"), stat = "identity", position = "identity") +
    geom_bar(aes(y = -Hypomethylation, fill = "Hypomethylation"), stat = "identity", position = "identity") +
    geom_text(aes(label = Hypermethylation, y = Hypermethylation), vjust = -0.5) +
    geom_text(aes(label = Hypomethylation, y = -Hypomethylation), vjust = 1.5) +
    scale_fill_manual(values = c("Hypermethylation" = "#FC8D62", "Hypomethylation" = "#66C2A5"),
                      breaks = c("Hypermethylation", "Hypomethylation"),
                      guide = guide_legend(title = "Methylation Status")) +
    labs(title = title) +
    theme_minimal() +
    theme(panel.grid.major.x = element_blank(),
          panel.border = element_rect(colour = "black", fill = NA),
          legend.text = element_text(size = 11),
          legend.title = element_text(size = 11),
          axis.text.x = element_text(angle = 90, hjust = 1),
          axis.text = element_text(size = 11),
          axis.text.y = element_blank(),   # Remove y-axis labels
          axis.title.y = element_blank()) +
    ylim(-max(plot_data$Hypomethylation)-space*1.5,max(plot_data$Hypermethylation)+space*1.5) +
    coord_fixed(ratio = bp_ratio)
}
```

```{r barplot-DMPs}
if(!params$skip_DM) {
  plot_data <- lapply(test_result_sig, function(x) table(as.vector(x[[3]]>0))) %>% do.call(rbind, .)
  colnames(plot_data) <- c("Hypomethylation","Hypermethylation")
  plot_data <- rownames_to_column(as.data.frame(plot_data), var = "Contrast")
  plot_data$Contrast <- factor(plot_data$Contrast, levels = names(test_result_sig))
  DM_barplots(plot_data, "Number of DMPs")
}
```

### Volcano plots of all CpGs

```{r volcano-function}
plot_volcano_dmp <- function(title, res, p.thres, fc.thres) {
  res <- res %>% as.data.frame()
  res_up <- res[res[, 3] > fc.thres & res[, 5] < p.thres, ] %>% na.omit()
  res_down <- res[res[, 3] < -fc.thres & res[, 5] < p.thres, ] %>% na.omit()
  # Prepare the base plot
  vplot <- ggplot(res, aes(.data[[names(res)[3]]], -log10(.data[[names(res)[5]]]), text = Probe_ID)) +
    geom_point(size = 2) +
    geom_hline(yintercept = -log10(p.thres), linetype = "longdash", colour = "grey", size = 1) +
    geom_vline(xintercept = fc.thres, linetype = "longdash", colour = "#BE684D", size = 1) +
    geom_vline(xintercept = -fc.thres, linetype = "longdash", colour = "#2C467A", size = 1) +
    labs(title = title, subtitle = "Volcano plot") +
    ylab("-log10 P") +
    xlab("DNA Methylation Difference") +
    theme_bw() +
    scale_x_continuous(limits = c(-1, 1), oob = scales::squish)
  
  # Highlight points based on thresholds
  if (nrow(res_up) > 0) { vplot <- vplot + geom_point(data = res_up, aes(x = .data[[names(res_up)[3]]], y = -log10(.data[[names(res_up)[5]]])), color = 'red') }
  if (nrow(res_down) > 0) { vplot <- vplot + geom_point(data = res_down, aes(x = .data[[names(res_down)[3]]], y = -log10(.data[[names(res_down)[5]]])), color = 'blue') }
  
  return(vplot)
}
plot_volcano_dmr <- function(title, res, p.thres, fc.thres) {
  res <- res %>% as.data.frame()
  res_up <- res[res[, "Seg_Est"] > fc.thres & res[, "Seg_Pval_adj"] < p.thres, ]
  res_down <- res[res[, "Seg_Est"] < -fc.thres & res[, "Seg_Pval_adj"] < p.thres, ]
  # Prepare the base plot
  vplot <- ggplot(res, aes(Seg_Est, -log10(Seg_Pval_adj), text = Seg_ID)) +
    geom_point(size = 2) +
    geom_hline(yintercept = -log10(p.thres), linetype = "longdash", colour = "grey", size = 1) +
    geom_vline(xintercept = fc.thres, linetype = "longdash", colour = "#BE684D", size = 1) +
    geom_vline(xintercept = -fc.thres, linetype = "longdash", colour = "#2C467A", size = 1) +
    labs(title = title, subtitle = "Volcano plot") +
    ylab("-log10 Pval Adj") +
    xlab("DNA Methylation Difference") +
    theme_bw() +
    scale_x_continuous(limits = c(-max(res[["Seg_Est"]]), max(res[["Seg_Est"]])), oob = scales::squish)
  
  # Highlight points based on thresholds
  if (nrow(res_up) > 0) { vplot <- vplot + geom_point(data = res_up, aes(Seg_Est, -log10(Seg_Pval_adj)), color = 'red') }
  if (nrow(res_down) > 0) { vplot <- vplot + geom_point(data = res_down, aes(Seg_Est, -log10(Seg_Pval_adj)), color = 'blue') }
  
  return(vplot)
}
```

```{r volcano}
if(!params$skip_DM) {
  for (x in names(test_result)) {
    vp_list_dmp <- plot_volcano_dmp(x, test_result[[x]], pval, eff_thres)
    print(vp_list_dmp)
  }
} 
```

## Differentially methylated regions (DMRs)

```{r warning flag, results='asis'}
if(!params$skip_DM & length(warning_flag)>0) {
  cat("***Note that for comparison(s) ", paste(warning_flag, sep = ", "), ", neither group had a replicate. So there was no p value could be produced for DMRs. The results below show all DMRs mapping to the CpGs with an effective size > ", eff_thres, ".***<br>", sep = "")
}
```

### Tables of significant DMRs

The tables below show unique significant DMRs. Non-chromosome CpGs were not shown. Please find the full significant DMR lists in the `DMR` folder.

```{r DMR tables}
if(!params$skip_DM) {
  htmltools::tagList(setNames(merged_dt, NULL))
}
```

### Barplots of significant DMRs

```{r barplot-DMRs}
if(!params$skip_DM) {
  plot_data <- lapply(merged_plt, function(x) table(factor(as.vector(x$Seg_Est > 0), levels = c(FALSE, TRUE)))) %>% do.call(rbind, .)
  colnames(plot_data) <- c("Hypomethylation","Hypermethylation")
  plot_data <- rownames_to_column(as.data.frame(plot_data), var = "Contrast")
  plot_data$Contrast <- factor(plot_data$Contrast, levels = names(merged_plt))
  DM_barplots(plot_data, "Number of DMRs")
}
```

### Volcano plots of all CpGs-merged regions

*Note that the number of significant DMR shown in the volcano plot can be larger than those in the bar-plots and tables above because non-chromosome CpGs are also shown here.*

```{r volcano regions}
if(!params$skip_DM) {
  for (x in names(merged)) {
    vp_list_dmr <- plot_volcano_dmr(x, merged[[x]], seg_fdr, eff_thres)
    print(vp_list_dmr)
  }
}
```

# Differentially methylated probe (DMP) enrichment 

```{r enrichment results folder}
if(!params$skip_DM) { 
  if(!dir.exists(file.path(params$result_folder, "Enrichment"))) dir.create(file.path(params$result_folder, "Enrichment"))
}
```

## Feature enrichment

Available database sets are listed in the table below. 

```{r database table}
if(!params$skip_DM) KYCG_listDBGroups(params$platform) %>% DT::datatable(rownames = FALSE)
```

### Significant DMPs enriched features

The tables below show features enriched with DMPs. Results were filtered with FDR < 0.05 and overlap > 10. 

```{r enrichment test}
if(!params$skip_DM) {
  if (!exists("query")) query <- lapply(test_result_sig, function(x) x$Probe_ID)
  if (!exists("cpg_enrichment_result")) cpg_enrichment_result <- lapply(query, function(x) testEnrichment(query = x, platform = params$platform))
  lapply(names(cpg_enrichment_result), function(x) cpg_enrichment_result[[x]] %>% dplyr::filter(overlap>10) %>% dplyr::filter(FDR<0.05) %>% write.csv(file = paste0(params$result_folder, "/Enrichment/DMP_feature_enrichment_", gsub("\\s", "_", x), ".csv"), row.names = FALSE))
  cpg_enrichment_result_dt <- lapply(names(cpg_enrichment_result), function(x) cpg_enrichment_result[[x]] %>% dplyr::filter(overlap>10) %>% dplyr::filter(FDR<0.05) %>% dplyr::select(dbname, estimate, test, FDR) %>% DT::datatable(caption = paste(x, "Feature Enrichments")))
  htmltools::tagList(setNames(cpg_enrichment_result_dt, NULL))
}
```

### Top 20 enriched features

The dot plots below show the top 20 enriched genes. Dot color indicates odds ratio (OR) in the Fisher's test. 

```{r enrichment dotplot, results = 'hide'}
if(!params$skip_DM) lapply(names(cpg_enrichment_result), function(x) KYCG_plotDot(cpg_enrichment_result[[x]], title = paste(x, "Top 20 Feature Enrichments"))) 
```

### Additional database sets

Additional GpG database sets available for the enrichment analysis are listed in the table below. This collection was downloaded from https://github.com/zhou-lab/KYCG_knowledgebase_EPICv2/tree/main. Please contact us further if you would like to test for the additional database. 

```{r additional databases}
if(!params$skip_DM) {
  more_dbs <- KYCG_loadDBs("KYCG_knowledgebase_EPICv2")
  lapply(more_dbs, attr, "group") %>% unlist %>% unique %>% as.tibble() %>% setNames("Database Group") %>% DT::datatable()
}
```

```{r additional enrichment test for selected database}
# test for a selected database (pending function ...)
# dbs <- KYCG_getDBs(group_nms = KYCG_listDBGroups(params$platform)[[1]][1], platform = params$platform)
# cpg_enrichment_result <- lapply(query, function(x) testEnrichment(x, databases = dbs, platform = params$platform))
```

## Gene enrichment

### Significant DMPs enriched genes

Test whether the DMP sets are enriched for any genes. Probe-gene distance for association was set 10000 at most. Results were filtered with FDR < 0.05 and overlap > 10. 

```{r gene enrichment}
if(!params$skip_DM) {
  if(!exists("gene_enrichment_result")) gene_enrichment_result <- lapply(query, function(x) testEnrichment(x, KYCG_buildGeneDBs(x, max_distance=10000), platform = params$platform))
  lapply(names(gene_enrichment_result), function(x) gene_enrichment_result[[x]] %>% dplyr::filter(overlap>10) %>% dplyr::filter(FDR<0.05) %>% write.csv(file = paste0(params$result_folder, "/Enrichment/DMP_gene_enrichment_", gsub("\\s", "_", x), ".csv"), row.names = FALSE))
  gene_enrichment_result_dt <- lapply(names(gene_enrichment_result), function(x) gene_enrichment_result[[x]] %>% dplyr::filter(overlap>10) %>% dplyr::filter(FDR<0.05) %>% dplyr::select(dbname, estimate, gene_name, FDR, nQ, nD, overlap) %>% DT::datatable(caption = paste(x, "Gene Enrichments")))
  htmltools::tagList(setNames(gene_enrichment_result_dt, NULL))
}
```

### Top 20 enriched genes

The dot plots below show the top 20 enriched genes. Dot color indicates odds ratio (OR) in the Fisher's test. 

```{r gene enrichment plot, results = 'hide'}
if(!params$skip_DM) lapply(names(gene_enrichment_result), function(x) KYCG_plotDot(gene_enrichment_result[[x]], label_by = "gene_name", title = paste(x, "Top 20 Gene Enrichments")))
```

## Gene-set over-representatoin analysis

Get DMP-associated genes and test whether these genes are enriched for any GO terms. Probe-gene distance for association was set 10000 at most. Results were filtered with FDR < 0.05. DMP-associated gene-sets with a size < 20 were excluded from the analysis. 

The dot plots below show the top 10 enriched GO terms or KEGG pathways. Please find the full results in tables in the `ORA` folder. 

```{r enrichment-dotplot-function}
gs_dotplot <- function(clusterProfResult, fdr_thres, title) {
  # get the top 10 enriched gene sets in each cluster to plot
  res2plot <- clusterProfResult@compareClusterResult %>% dplyr::filter(p.adjust < fdr_thres) %>% dplyr::group_by(Cluster) %>% dplyr::slice_head(n = 10) %>% dplyr::ungroup()
  if (nrow(res2plot)==0) { return(NULL) }
  # calculate gene ratio
  res2plot$GeneRatio <- sapply(strsplit(res2plot$GeneRatio, "/"), function(x) as.numeric(x[1]) / as.numeric(x[2]))
  
  # set the levels of x and y axis
  res2plot$Cluster <- factor(res2plot$Cluster, levels = names(clusterProfResult@geneClusters))
  res2plot$Description <- factor(res2plot$Description, levels = unique(res2plot$Description))
  
  p <- ggplot(res2plot, aes(x = Cluster, y = Description)) +
    geom_point(aes(size = GeneRatio, colour = p.adjust), show.legend = T) + 
    theme_bw() +
    theme(
      axis.text = element_text(colour = "black"),
      axis.text.x = element_text(size = 12, angle = 90, vjust = 0.5, hjust = 1),
      axis.text.y = element_text(size = 10),
      legend.title = element_text(size = 14),
      legend.text = element_text(size = 12),
      plot.margin = unit(c(0.5, 0.5, 0.1, 0.1), "cm")
    ) +
    labs(title = title, x = "", y = "", caption = "") +
    scale_colour_gradient(name = "FDR adjusted\n p-value\n", limits = c(0, fdr_thres)) +
    scale_size_continuous(name = "Gene Ratio")
  print(p)
}
```

```{r gene input}
if(!params$skip_DM) {
  if(!exists("ora_genes")) ora_genes <- lapply(query, function(x) lapply(KYCG_buildGeneDBs(x, max_distance = 10000), function(y) attributes(y)$gene_name) %>% unlist(use.names = FALSE) %>% unique())
}
```

```{r functional enrichment, fig.height=12, fig.width=12}
if(!params$skip_DM) {
  if (params$species == "human") {
    gsorDb <- "org.Hs.eg.db"
    keggOrg <- 'hsa'
    msigdbr_species <- "Homo sapiens"
  } else if (params$species == "mouse") {
    gsorDb <- "org.Mm.eg.db"
    keggOrg <- 'mmu'
    msigdbr_species <- "Mus musculus"
  } else if (params$species == "rat") {
    gsorDb <- "org.Rn.eg.db"
    keggOrg <- 'rno'
    msigdbr_species <- "Rattus norvegicus"
  } else {
    stop("Species not supported. ")
  }

  deGenes_ora <- ora_genes[sapply(ora_genes, length) >= 20]

  if (!exists("enrichGO_res")) {
    enrichGO_res <- list()
    go_cat <- c('BP','CC','MF')

    if (!all(sapply(deGenes_ora, length) == 0)) {
      # run GO enrichment analysis
      enrichGO_res$all <- lapply(go_cat, function(x) clusterProfiler::compareCluster(geneCluster = deGenes_ora, fun = 'enrichGO', keyType = 'SYMBOL', OrgDb = gsorDb, ont = x, readable = TRUE)) %>% setNames(go_cat)
      # remove null elements
      enrichGO_res$all <- if (all(sapply(enrichGO_res$all, is.null))) NULL else enrichGO_res$all[!sapply(enrichGO_res$all, is.null)]
    }
  }

  if (!is.null(enrichGO_res$all)) {
    # plot the enriched GO terms
    lapply(names(enrichGO_res$all), function(x) gs_dotplot(enrichGO_res$all[[x]], params$ora_fdr_thres, paste0("Top 10 enriched GO ", x, " terms per contrast"))) %>% invisible()
    # save the enriched GO terms
    lapply(names(enrichGO_res$all), function(x) lapply(names(deGenes_ora), function(y) readr::write_csv(subset(enrichGO_res$all[[x]]@compareClusterResult, Cluster==y & p.adjust<params$ora_fdr_thres), file = paste0(file.path(params$result_folder, "Enrichment"), "/ORA_GO_", x, "_", gsub(" ", "_", y), "_padj", sub("0\\.", "", as.character(params$ora_fdr_thres)), ".csv")))) %>% invisible()
  }

  if (!exists("enrichKEGG_res")) {
    enrichKEGG_res <- list()

    if (!all(sapply(deGenes_ora, length) == 0)) {
      # map gene names to entrez ids
      deEntrez <- lapply(deGenes_ora, function(x) bitr(x, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = gsorDb)$ENTREZID)
      # run KEGG enrichment analysis
      enrichKEGG_res$all <- clusterProfiler::compareCluster(geneCluster = deEntrez, fun = 'enrichKEGG', organism = keggOrg)
      # map entrez ids back to gene names
      enrichKEGG_res$all@compareClusterResult$geneID <- lapply(enrichKEGG_res$all@compareClusterResult$geneID, function(x) unlist(strsplit(x, "/")) %>% bitr(fromType = "ENTREZID", toType = "SYMBOL", OrgDb = gsorDb) %>% dplyr::select(SYMBOL) %>% unlist() %>% paste(collapse = "/")) %>% unlist()
    }
  }

  if (!is.null(enrichKEGG_res$all)) {
    # plot the enriched KEGG pathways
    gs_dotplot(enrichKEGG_res$all, params$ora_fdr_thres, "Top 10 enriched KEGG pathways per contrast")
    # save the enriched KEGG pathways
    lapply(names(deGenes_ora), function(x) readr::write_csv(subset(enrichKEGG_res$all@compareClusterResult, Cluster==x & p.adjust<params$ora_fdr_thres), file = paste0(file.path(params$result_folder, "Enrichment"), "/ORA_KEGG_", x, "_padj", sub("0\\.", "", as.character(params$ora_fdr_thres)), ".csv"))) %>% invisible()
  }
}
```

# Metagene methylation levels

```{r metagene}
betas_meta <- betas
rownames(betas_meta) <- gsub("_.*", "", rownames(betas_meta))
KYCG_plotMeta(betas_meta)
```

# Track view with heatmap

We are able to visualize the methylation level across a given genomic region, around a given gene, or around a given probe. Below shows an example of the visualization. Please contact us further if you would like to visualize certain genomic regions. 

```{r track view with heatmap}
visualizeGene('DNMT1', betas)
# visualizeRegion('chr15', 96343814, 96544301, betas, platform="EPIC", show.probeNames = FALSE)
# visualizeGene('DNMT1', betas, platform="EPIC", genome = "hg38")
# visualizeProbes(c("cg00000165"), betas, platform="EPIC", genome = "hg38", upstream = 2000, dwstream = 2000)
```

```{r save enrichment results}
if(!params$skip_DM && !params$test_mode) save(list = ls()[-grep("params", ls())], file = paste0(params$result_folder, ".RData"))
```

------------------------------------------------------------------------

# Session info

The output from running 'sessionInfo' is shown below and details all packages and version necessary to reproduce the results in this report.

```{r session info}
sessionInfo()
```
