---
title: "CRI-BIO-XXX: Demostration"
author: "CRI Bioinformatics Core"
date: "_`r Sys.Date()`_"
output:
  html_document:
    code_folding: show
    df_print: paged
    theme: yeti
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    number_sections: true
    self_contained: true
params:
  raw_data_root_folder: raw_data
  sampleinfo: sampleinfo.csv
  platform: EPICv2
  result_folder: Demostration
  remove_bad_samples: true
  unmask: false
  extra_mask: false
  pOOBAH_pval: 0.05
  match_design: true
  batch_correction_method: combat
  batch_to_correct: "2025-09-25"
  tSNE_perplexity: 12
  skip_DM: false
  test_mode: true
  contrast_pairs: !expr list(
      c("Group1", "Group2"),
      c("Group1", "Group3")
    )
  species: human
  ora_fdr_thres: 0.05
---

```{r setup, include=FALSE}

library(ggplot2)
library(plotly)
library(patchwork)
library(Rtsne)
library(matrixStats)
library(dplyr)
library(tibble)
library(stats)
library(circlize)
library(dendextend)
library(VIM)
library(sesame)
library(ComplexHeatmap)
library(ggrepel)
library(org.Hs.eg.db)
library(org.Mm.eg.db)
library(clusterProfiler)
library(limma)
library(sva)
library(RPMM)
library(minfi)
library(AnnotationHub)
library(DNAcopy)
library(knowYourCG)

sesameDataCache()

knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

if(!dir.exists(params$result_folder)) dir.create(params$result_folder)

```

# Experiment design

We use [sesame](https://bioconductor.org/packages/release/bioc/vignettes/sesame/inst/doc/sesame.html) to analyze the data as recommended by Illumina.

We use the standard openSesame workflow to process raw signal data to beta values. A description of the method details can be found in [this article](https://epicom.biomedcentral.com/articles/10.1186/s43682-023-00021-5). 

The following steps were executed during preprocessing: 

- 1) probes of poor design were masked, including the recommended probes to mask by Sesame, and probes that are cross reactive with other parts of the genome, which is indicated by the CH_WGBS_evidence column in [EPICv2manifest](https://bioconductor.statistik.tu-dortmund.de/packages/3.19/data/annotation/vignettes/EPICv2manifest/inst/doc/EPICv2manifest.html) (when `extra_mask` is `TRUE`).
- 2) color channels for Type-I probes were inferred and reset instead of using what is specified in manifest.
- 3) dye bias was corrected using the nonlinear scaling method in Sesame to address the intensity differences between green and red channels.
- 4) background subtraction: background was modeled in a normal distribution and true signal in an exponential distribution, for which the deconvolution is parameterized using Out-Of-Band (oob) probes.
- 5) experiment-dependent masking based on signal detection p-values is performed to exclude artifactual methylation-level reading and probes with too much influence from signal background.

```{r pre-processing}

if (file.exists(paste0(params$result_folder, ".RData"))){
  cat(paste0(params$result_folder, ".RData"), " already exists in the working directory. Use it for analyses downstream of QC instead of reading from raw data. If you want to restart from reading in raw data, please remove ", paste0(params$result_folder, ".RData"), " from current working directory and knit this script again. ")
  load(paste0(params$result_folder, ".RData"))
}else{
  metadata = list.files(params$raw_data_root_folder, pattern = ".csv", full.names = T, recursive = T)
  metadata = lapply(metadata, function(x) read.csv(x, skip = 5))
  metadata = do.call(rbind, metadata)

  sampleinfo = read.csv(params$sampleinfo, na.strings = "")

  coldata = merge(metadata, sampleinfo, by = "Sample_Name")
  coldata$IDAT = paste(coldata$Sentrix_ID, coldata$Sentrix_Position, sep = "_")
  coldata = coldata[,c("IDAT", "Sample_Name", names(sampleinfo)[-1])]

  if(nrow(metadata)>nrow(coldata))  cat("There are ", nrow(metadata), " samples found in the raw data folder, but only ", nrow(coldata), " samples have experimental condition records. We will omit samples without any experimental condition record from the downstream analysis. ")

  idats = searchIDATprefixes(params$raw_data_root_folder)
  
  sdfs = lapply(idats, readIDATpair, platform = params$platform) %>% setNames(gsub(".*/", "", idats))
  sdfs = sdfs[coldata$IDAT]
  
  ah <- AnnotationHub()
  EPICv2manifest <- ah[["AH116484"]]

  # listAvailableMasks("EPICv2") to check available masks
  # By default, the recommended masks for EPICv2 are:
  # c("M_mapping","M_nonuniq","M_1baseSwitchSNPcommon_5pt","M_2extBase_SNPcommon_5pt","M_SNPcommon_5pt")
  sdfs_pp = openSesame(sdfs, prep = "QCDPB", 
                       prep_args = list(Q=list(mask_names = "recommended"), P=list(pval.threshold = params$pOOBAH_pval)), 
                       func = NULL)
  if (params$extra_mask) {
    sdfs_pp = lapply(sdfs_pp, addMask, probes = rownames(EPICv2manifest)[EPICv2manifest[["CH_WGBS_evidence"]]=="Y"])
  }
  if (params$match_design) {
    sdfs_pp = lapply(sdfs_pp, matchDesign)
  }
  
  betas = lapply(sdfs_pp, getBetas, collapseToPfx=FALSE) %>% do.call(cbind, .)
  betas_unmask = lapply(sdfs_pp, getBetas, collapseToPfx=FALSE, mask=FALSE) %>% do.call(cbind, .)
  coldata = coldata[na.omit(match(colnames(betas), coldata$IDAT)),]
  
  segs <- lapply(sdfs_pp, function(x) cnSegmentation(x))
}

```

# Experiment design

## Sample information for each IDAT pair

```{r table1}
if (file.exists(paste0(params$result_folder, ".RData"))){
  sample_dt
}else{
  sample_dt = DT::datatable(coldata, rownames = FALSE)
  sample_dt
}

```

## Sample counts per condition

```{r table2}
if (file.exists(paste0(params$result_folder, ".RData"))){
  freq_dt
}else{
  freq_dt = coldata[, -c(1, 2), drop = F] %>% group_by(across(everything())) %>% summarise(Frequency = n()) %>% ungroup() %>%  as.data.frame() %>% DT::datatable(rownames = FALSE)
  freq_dt
}
```

# QC

## Bisulfite conversion efficiency

The closer GCT score to 1, the more effecient the bisulfite conversion is.

```{r bisConversionControl}
mft = sesameAnno_buildManifestGRanges(sprintf(
    "%s/EPICv2/EPICv2.hg38.manifest.tsv.gz",
    "https://github.com/zhou-lab/InfiniumAnnotationV1/raw/main/Anno/"),
    columns="nextBase")
extR = names(mft)[!is.na(mft$nextBase) & mft$nextBase=="R"]
extA = names(mft)[!is.na(mft$nextBase) & mft$nextBase=="A"]

lapply(sdfs, function(x) bisConversionControl(x, extR = extR, extA = extA)) %>% unlist() %>% as.data.frame() %>% setNames("GCT score") %>% DT::datatable(rownames = TRUE)

```

## Tables (Before pre-processing)

```{r QC}

if (!file.exists(paste0(params$result_folder, ".RData"))){
  qcs = list()
  qcs_df = list()
  qcs_dt = list()
  # rankstats = list()
  # rankstats_df = list()
  # rankstats_dt = list()
  for (f in c("detection", "intensity", "numProbes", "channel", "dyeBias", "betas")) {
    qcs[[f]] = openSesame(sdfs, prep="", func=sesameQC_calcStats, funs=f)
    qcs_df[[f]] = do.call(rbind, lapply(qcs[[f]], as.data.frame))
    colnames(qcs_df[[f]]) = qcs[[f]][[1]]@group[[1]][colnames(qcs_df[[f]])]
    qcs_dt[[f]] = DT::datatable(data = qcs_df[[f]], caption = names(qcs[[f]][[1]]@group), rownames = T)
    # rankstats[[f]] = lapply(qcs[[f]], sesameQC_rankStats, platform=params$platform_plt)
    # rankstats_df[[f]] = do.call(rbind, lapply(rankstats[[f]], as.data.frame))
    # colnames(rankstats_df[[f]]) = rankstats[[f]][[1]]@group[[1]][colnames(rankstats_df[[f]])]
    # rankstats_dt[[f]] = DT::datatable(data = rankstats_df[[f]], caption = names(rankstats[[f]][[1]]@group), rownames = T)
  }
}

htmltools::tagList(setNames(qcs_dt, NULL))

```

## Selected barplots (Before pre-processing)

Please find the full set of barplots in the `QC-Bar` folder.

```{r barplots }

if (!file.exists(paste0(params$result_folder, ".RData"))){
  if(!dir.exists(paste0(params$result_folder, "/QC-Bar"))) dir.create(paste0(params$result_folder, "/QC-Bar"))

  for (f in c("detection", "intensity", "numProbes", "channel", "dyeBias", "betas")) {
    h = length(qcs[[f]][[1]]@group[[1]])*250
    png(paste0(params$result_folder, "/QC-Bar/barplot_", f, "_before.png"), width = 500, height = h, units = "px")
    sesameQC_plotBar(qcs[[f]], keys = "") %>% print()
    dev.off()
  }
}

for (f in c("detection", "intensity", "dyeBias", "betas")) {
  sesameQC_plotBar(qcs[[f]]) %>% print()
}

```

## Tables (After pre-processing)

```{r QC 2}

if (!file.exists(paste0(params$result_folder, ".RData"))){
  qcs_pp = list()
  qcs_pp_df = list()
  qcs_pp_dt = list()
  for (f in c("detection", "intensity", "numProbes", "channel", "dyeBias", "betas")) {
    qcs_pp[[f]] = openSesame(sdfs_pp, prep="", func=sesameQC_calcStats, funs=f)
    qcs_pp_df[[f]] = do.call(rbind, lapply(qcs_pp[[f]], as.data.frame))
    colnames(qcs_pp_df[[f]]) = qcs_pp[[f]][[1]]@group[[1]][colnames(qcs_pp_df[[f]])]
    qcs_pp_dt[[f]] = DT::datatable(data = qcs_pp_df[[f]], caption = names(qcs_pp[[f]][[1]]@group), rownames = T)
  }
}

htmltools::tagList(setNames(qcs_pp_dt, NULL))

```

## Selected barplots (After pre-processing)

Please find the full set of barplots in the `QC-Bar` folder.

```{r barplots 2}

if (!file.exists(paste0(params$result_folder, ".RData"))){
  if(!dir.exists(paste0(params$result_folder, "/QC-Bar"))) dir.create(paste0(params$result_folder, "/QC-Bar"))

  for (f in c("detection", "intensity", "numProbes", "channel", "dyeBias", "betas")) {
    h = length(qcs_pp[[f]][[1]]@group[[1]])*250
    png(paste0(params$result_folder, "/QC-Bar/barplot_", f, "_after.png"), width = 500, height = h, units = "px")
    sesameQC_plotBar(qcs_pp[[f]], keys = "") %>% print()
    dev.off()
  }
}

for (f in c("detection", "intensity", "dyeBias", "betas")) {
  sesameQC_plotBar(qcs_pp[[f]]) %>% print()
}

```

```{r BetaByDesign, include=FALSE }

if (!file.exists(paste0(params$result_folder, ".RData"))){
  if(!dir.exists(paste0(params$result_folder, "/QC-BetaByDesign"))) dir.create(paste0(params$result_folder, "/QC-BetaByDesign"))

  for (f in names(sdfs)) {
    png(paste0(params$result_folder, "/QC-BetaByDesign/densityplot_", f, ".png"), width = 500, height = 500, units = "px")
    par(mfrow=c(2,1), mar=c(3,3,2,1))
    sesameQC_plotBetaByDesign(sdfs[[f]], main="Before Pre-processing", xlab="\beta") %>% print()
    sesameQC_plotBetaByDesign(sdfs_pp[[f]], main="After Pre-processing", xlab="\beta") %>% print()
    dev.off()
  }
}

```

```{r RedGrnQQ, include=FALSE }

if (!file.exists(paste0(params$result_folder, ".RData"))){
  if(!dir.exists(paste0(params$result_folder, "/QC-RedGrnQQ"))) dir.create(paste0(params$result_folder, "/QC-RedGrnQQ"))

  for (f in names(sdfs)) {
    png(paste0(params$result_folder, "/QC-RedGrnQQ/qqplot_", f, ".png"), width = 900, height = 500, units = "px")
    par(mfrow=c(1,2), mar=c(3,3,2,1))
    sesameQC_plotRedGrnQQ(sdfs[[f]], main = "Before Pre-processing") %>% print()
    sesameQC_plotRedGrnQQ(sdfs_pp[[f]], main = "After Pre-processing") %>% print()
    dev.off()
  }
}

```

```{r IntensVsBetas, include=FALSE }

if (!file.exists(paste0(params$result_folder, ".RData"))){
  if(!dir.exists(paste0(params$result_folder, "/QC-IntensVsBetas"))) dir.create(paste0(params$result_folder, "/QC-IntensVsBetas"))

  for (f in names(sdfs)) {
    png(paste0(params$result_folder, "/QC-IntensVsBetas/intbetaplot_", f, ".png"), width = 900, height = 500, units = "px")
    par(mfrow=c(1,2), mar=c(3,3,2,1))
    sesameQC_plotIntensVsBetas(sdfs[[f]]) %>% print()
    sesameQC_plotIntensVsBetas(sdfs_pp[[f]]) %>% print()
    dev.off()
  }
}

```

## Sample counts per condition after removing bad samples

```{r table3, results='asis'}

if(params$remove_bad_samples){
  samples_rm <- rownames(qcs_df$detection)[qcs_df$detection[grep("% Detection Success\\s+$",names(qcs_df$detection))]<0.4]
  
  if (length(samples_rm)>0) {
    coldata <- coldata[!coldata$IDAT %in% samples_rm,]
    betas <- betas[,!colnames(betas) %in% samples_rm]
    betas_unmask <- betas_unmask[,!colnames(betas_unmask) %in% samples_rm]
    cat("Bad samples (detection success rate < 40%) to remove:", paste(samples_rm, collapse = ", "))
    
    coldata[, -c(1, 2), drop = F] %>% group_by(across(everything())) %>% summarise(Frequency = n()) %>% ungroup() %>%  as.data.frame() %>% DT::datatable(rownames = FALSE)
  }else{
    cat("There is no bad sample. All samples have a detection success rate >= 40%.")
  }
}else{
  samples_rm <- rownames(qcs_df$detection)[qcs_df$detection[grep("% Detection Success\\s+$",names(qcs_df$detection))]<0.4]
  if (length(samples_rm)>0) {
    cat("Potential bad samples (detection success rate < 40%) to remove:", paste(samples_rm, collapse = ", "), "<br>")
    cat("For analyses in this report, we use all samples. ")
  }else{
    cat("There is no bad sample. All samples have a detection success rate >= 40%.")
  }
}

```

```{r save env}
if (!file.exists(paste0(params$result_folder, ".RData"))) save(list = ls()[-grep("params", ls())], file = paste0(params$result_folder, ".RData"))
```

## Batch correction

```{r batch correction, results='asis'}
if(params$unmask){
  betas_clean <- na.omit(betas_unmask)
}else{
  betas_clean <- na.omit(betas)
}
if(!is.null(params$batch_correction_method)){
  if(params$batch_correction_method=="limma"){
    coldata$Batch <- factor(coldata$Batch)
    coldata$Group <- factor(coldata$Group)
  
    # design matrix must be numeric
    design <- model.matrix(~ Group, data = coldata)
  
    # Run batch correction
    betas_clean <- limma::removeBatchEffect(
        betas_clean,
        batch = coldata$Batch,
        design = design
    )
    cat("Note that batch effect correction has been done using limma::removeBatchEffect\n")
  }else if(params$batch_correction_method=="combat"){
    mod <- model.matrix(~ Group, coldata)
    betas_clean <- sva::ComBat(betas_clean, batch=coldata$Batch, mod=mod)
    cat("Note that batch effect correction has been done using ComBat.\n")
  }else if(params$batch_correction_method=="icombat"){
    source(file = "iComBat-main/scripts/icombat.R")
    betas_clean_init <- ComBatInitialize(
      betas_clean[,!coldata$Batch%in%params$batch_to_correct],
      coldata$Batch[!coldata$Batch%in%params$batch_to_correct],
      mod = NULL,
      mean.only = FALSE
    )
    betas_clean[,coldata$Batch%in%params$batch_to_correct] <- ComBatIncremental(
      betas_clean[,coldata$Batch%in%params$batch_to_correct],
      params$batch_to_correct,
      betas_clean_init,
      mod.new = NULL,
      par.prior = TRUE,
      prior.plots = FALSE,
      intercept.col = 1
    )
    cat("Note that batch effect correction has been done using iComBat.\n")
  }else{
    stop("Please provide one of the following batch correction methods: limma, combat, icombat.")
  }
}else{
  cat("Note that batch effect correction was skipped.\n")
}
```

<!-- ```{r remove suspecious sample 1, results='asis'}
coldata <- coldata[!coldata$IDAT=="208781040014_R05C01",]
betas <- betas[,!colnames(betas)=="208781040014_R05C01"]
betas_unmask <- betas_unmask[,!colnames(betas_unmask)=="208781040014_R05C01"]
betas_clean <- betas_clean[,!colnames(betas_clean)=="208781040014_R05C01"]
segs[["208781040014_R05C01"]] <- NULL
cat("Suspecious FATWO sample 208781040014_R05C01 clustered to Endometrioid Carcinoma. Remove it temporarily. ")
``` -->

# t-SNE plots

For the t-SNE analysis below, we applied perplexity = `r params$tSNE_perplexity`. 

```{r t-SNE notes, results='asis'}
if(params$unmask){
  cat("Note that we used a probe matrix without masking bad detection p-values for the t-SNE analysis. ", nrow(betas_unmask), " CpG loci were used here.")
}else{
  cat("Note that probes with bad detection p-values were masked as NAs in the QC steps. These probes were removed in the t-SNE analysis since they were unacceptable as inputs for t-SNE. ", nrow(na.omit(betas)), " out of ", nrow(betas), " CpG loci were removed here.")
}
```

## Using all CpG loci (probes)

```{r t-SNE function, include=FALSE}

tsne_plot <- function(tsne_result, coldata, title){
  tsne_data <- data.frame(
    tSNE1 = tsne_result$Y[,1],
    tSNE2 = tsne_result$Y[,2]
  )
  plot_data <- cbind(tsne_data, coldata[,-1])
  if (ncol(plot_data)==4) {
    gp <- ggplot(plot_data, aes(x = tSNE1, y = tSNE2, label = Sample_Name, color = !!sym(names(plot_data)[4]))) +
      geom_point(size = 2) +
      labs(title = title) +
      theme_minimal() +
      scale_color_brewer(palette = "Set2")
  }else if (ncol(plot_data)>4) {
    gp <- ggplot(plot_data, aes(x = tSNE1, y = tSNE2, label = Sample_Name, color = !!sym(names(plot_data)[4]), shape = !!sym(names(plot_data)[5]))) +
      geom_point(size = 2) +
      labs(title = title) +
      theme_minimal() +
      scale_color_brewer(palette = "Set2")
  }else{
    gp <- ggplot(plot_data, aes(x = tSNE1, y = tSNE2, label = Sample_Name)) +
      geom_point(size = 2) +
      labs(title = title) +
      theme_minimal() +
      scale_color_brewer(palette = "Set2")
  }
  return(gp)
}

```

```{r t-SNE all, include=FALSE}

set.seed(123)  # Set a seed for reproducibility
tsne_result <- Rtsne(t(betas_clean), dims = 2, perplexity = params$tSNE_perplexity, verbose = TRUE)
gp <- tsne_plot(tsne_result, coldata, "t-SNE clustering")

```

```{r plot out all, fig.width=8, results='asis'}
ggplotly(gp, tooltip = "label")
```

## Using the top 10k most variable CpG loci (probes)

```{r t-SNE top10k, include=FALSE}

top10k <- rownames(betas_clean)[order(rowVars(betas_clean), decreasing=TRUE)] %>% head(10000)
betas_clean_top <- betas_clean[top10k,]

set.seed(123)  # Set a seed for reproducibility
tsne_result <- Rtsne(t(betas_clean_top), dims = 2, perplexity = params$tSNE_perplexity, verbose = TRUE)

gp <- tsne_plot(tsne_result, coldata, "t-SNE clustering")

```

```{r plot out top10k, fig.width=8, results='asis'}
ggplotly(gp, tooltip = "label")
```

## Using the top 1k most variable CpG loci (probes)

```{r t-SNE top1k, include=FALSE}

top1k <- rownames(betas_clean)[order(rowVars(betas_clean), decreasing=TRUE)] %>% head(1000)
betas_clean_top <- betas_clean[top1k,]

set.seed(123)  # Set a seed for reproducibility
tsne_result <- Rtsne(t(betas_clean_top), dims = 2, perplexity = params$tSNE_perplexity, verbose = TRUE)

gp <- tsne_plot(tsne_result, coldata, "t-SNE clustering")

```

```{r plot out top1k, fig.width=8, results='asis'}
ggplotly(gp, tooltip = "label")
```

# MDS plots

MDS plot is better for showing global distance and evaluating before/after batch effects correction. 

```{r MDS notes, results='asis'}
if(params$unmask){
  cat("Note that we used a probe matrix without masking bad detection p-values for the MDS analysis. ", nrow(betas_unmask), " CpG loci were used here.")
}else{
  cat("Note that probes with bad detection p-values were masked as NAs in the QC steps. These probes were removed from the MDS analysis. ", nrow(na.omit(betas)), " out of ", nrow(betas), " CpG loci were removed here.")
}
```

```{r MDS function, include=FALSE}

mds_plot <- function(betas_matrix, coldata, title = "MDS Plot") {

  ## 1. Compute distance + classical MDS
  d <- dist(t(betas_matrix))              # distance across samples
  mds <- cmdscale(d, k = 2)               # extract 2 dimensions
  mds_df <- as.data.frame(mds)
  colnames(mds_df) <- c("MDS1", "MDS2")

  ## 2. Bind metadata (assuming coldata row order matches samples)
  plot_data <- cbind(mds_df, coldata[,-1])  # remove IDAT column

  ## 3. Plot
  if (ncol(plot_data) == 4) {
    gp <- ggplot(plot_data,
                 aes(x = MDS1, y = MDS2,
                     label = Sample_Name,
                     color = !!sym(names(plot_data)[4]))) +
      geom_point(size = 3) +
      labs(title = title) +
      theme_minimal() +
      scale_color_brewer(palette = "Set2")

  } else if (ncol(plot_data) > 4) {
    gp <- ggplot(plot_data,
                 aes(x = MDS1, y = MDS2,
                     label = Sample_Name,
                     color = !!sym(names(plot_data)[4]),
                     shape = !!sym(names(plot_data)[5]))) +
      geom_point(size = 3) +
      labs(title = title) +
      theme_minimal() +
      scale_color_brewer(palette = "Set2")
  }

  return(gp)
}

```

## Using all CpG loci (probes)

```{r MDS all, include=FALSE}
gp <- mds_plot(betas_clean, coldata, "MDS Plot")
```

```{r plot out MDS all, fig.width=8, results='asis'}
ggplotly(gp, tooltip = "label")
```

## Using the top 10k most variable CpG loci (probes)

```{r MDS top10k, include=FALSE}

top10k <- rownames(betas_clean)[order(rowVars(betas_clean), decreasing=TRUE)] %>% head(10000)
betas_clean_top <- betas_clean[top10k,]
gp <- mds_plot(betas_clean_top, coldata)

```

```{r plot out MDS top10k, fig.width=8, results='asis'}
ggplotly(gp, tooltip = "label")
```

## Using the top 1k most variable CpG loci (probes)

```{r MDS top1k, include=FALSE}

top1k <- rownames(betas_clean)[order(rowVars(betas_clean), decreasing=TRUE)] %>% head(1000)
betas_clean_top <- betas_clean[top1k,]
gp <- mds_plot(betas_clean_top, coldata)

```

```{r plot out MDS top1k, fig.width=8, results='asis'}
ggplotly(gp, tooltip = "label")
```

# Heatmap with unsupervised hierarchical clustering

## Top10k probes

```{r heatmap notes, results='asis'}
if(params$unmask){
  cat("Note that we used a probe matrix without masking bad detection p-values for the heatmap and unsupervised hierarchical clustering (Ward’s method). The unsupervised hierarchical clustering was done using the top 10k probes with the highest Median Absolute Deviation (MAD) whose beta values were shown on the heatmap. Probes were also clustered with the Ward's method. ")
}else{
  cat("The unsupervised hierarchical clustering (Ward’s method) was done using the top 10k probes with the highest Median Absolute Deviation (MAD) whose beta values were shown on the heatmap. Some beta values were masked due to low confidence in the detection of signals. It’s usually not recommended to use these low-confidence values in the downstream DM analysis. For heatmap visualization purpose, we removed any probes containing an NA value. Probes were then clustered with the Ward's method. ")
}
```

```{r heatmap save, include=FALSE }

if(params$unmask){
  if(is.null(params$batch_correction_method)){
    mad_values <- apply(betas_unmask, 1, mad)  # Calculate MAD, ignoring NAs
    betas.sel <- betas_unmask[order(mad_values, decreasing = TRUE)[1:10000], ]  # Subset the data to include only the top probes
  }
}else{
  mad_values <- apply(betas_clean, 1, mad, na.rm = TRUE)  # Calculate MAD, ignoring NAs
  betas.sel <- betas_clean[order(mad_values, decreasing = TRUE)[1:10000], ]  # Subset the data to include only the top probes
  # betas.sel <- imputeBetas(betas.sel) # CANNOT do imputation since it turned out to alter the relationship between samples.
  betas.sel <- na.omit(betas.sel)
}

sample_anno <- coldata[,-c(1,2), drop = FALSE]
sample_anno[] = lapply(sample_anno, as.factor)
annotation_colors <- lapply(sample_anno, function(x) {
  colors <- c(RColorBrewer::brewer.pal(8, "Set2"), RColorBrewer::brewer.pal(9, "Set1"))[1:length(levels(x))]
  setNames(colors[seq_along(levels(x))], levels(x))
})

pdf(file.path(params$result_folder, "heatmap_top10k.pdf"), width = 8, height = 16)
hm <- Heatmap(
  betas.sel,
  name = "Beta Values",  
  col = colorRamp2(c(0, 0.5, 1), c("blue", "whitesmoke", "red")),  
  na_col = "black",  
  show_row_names = FALSE,
  cluster_rows = TRUE,
  clustering_method_columns = "ward.D2",  # Specify clustering method for columns
  clustering_method_rows = "ward.D2",
  use_raster = TRUE,
  top_annotation = HeatmapAnnotation(
    df = sample_anno,
    col = annotation_colors,  # Assign colors automatically generated for each factor
    annotation_name_side = "left"  # Position of the annotation name
  )
)
draw(hm)
dev.off()

```

```{r heatmap plot, results='asis'}
cat(paste0('<iframe src="', file.path(params$result_folder, "heatmap_top10k.pdf"), '" width="100%" height="600px"></iframe>'))
```

## Top1k probes

```{r heatmap notes 2, results='asis'}
if(params$unmask){
  cat("Note that we used a probe matrix without masking bad detection p-values for the heatmap and unsupervised hierarchical clustering (Ward’s method). The unsupervised hierarchical clustering was done using the top 1k probes with the highest Median Absolute Deviation (MAD) whose beta values were shown on the heatmap. Probes were also clustered with the Ward's method. ")
}else{
  cat("The unsupervised hierarchical clustering (Ward’s method) was done using the top 1k probes with the highest Median Absolute Deviation (MAD) whose beta values were shown on the heatmap. Some beta values were masked due to low confidence in the detection of signals. It’s usually not recommended to use these low-confidence values in the downstream DM analysis. For heatmap visualization purpose, we removed any probes containing an NA value. Probes were then clustered with the Ward's method. ")
}
```

```{r heatmap save 2, include=FALSE }

if(params$unmask){
  if(is.null(params$batch_correction_method)){
    mad_values <- apply(betas_unmask, 1, mad)  # Calculate MAD, ignoring NAs
    betas.sel <- betas_unmask[order(mad_values, decreasing = TRUE)[1:1000], ]  # Subset the data to include only the top probes
  }
}else{
  mad_values <- apply(betas_clean, 1, mad, na.rm = TRUE)  # Calculate MAD, ignoring NAs
  betas.sel <- betas_clean[order(mad_values, decreasing = TRUE)[1:1000], ]  # Subset the data to include only the top probes
  # betas.sel <- imputeBetas(betas.sel) # CANNOT do imputation since it turned out to alter the relationship between samples.
  betas.sel <- na.omit(betas.sel)
}

sample_anno <- coldata[,-c(1,2), drop = FALSE]
sample_anno[] = lapply(sample_anno, as.factor)
annotation_colors <- lapply(sample_anno, function(x) {
  colors <- c(RColorBrewer::brewer.pal(8, "Set2"), RColorBrewer::brewer.pal(9, "Set1"))[1:length(levels(x))]
  setNames(colors[seq_along(levels(x))], levels(x))
})

pdf(file.path(params$result_folder, "heatmap_top1k.pdf"), width = 8, height = 16)
hm <- Heatmap(
  betas.sel,
  name = "Beta Values",  
  col = colorRamp2(c(0, 0.5, 1), c("blue", "whitesmoke", "red")),  
  na_col = "black",  
  show_row_names = FALSE,
  cluster_rows = TRUE,
  clustering_method_columns = "ward.D2",  # Specify clustering method for columns
  clustering_method_rows = "ward.D2",
  use_raster = TRUE,
  top_annotation = HeatmapAnnotation(
    df = sample_anno,
    col = annotation_colors,  # Assign colors automatically generated for each factor
    annotation_name_side = "left"  # Position of the annotation name
  )
)
draw(hm)
dev.off()

```

```{r heatmap plot 2, results='asis'}
cat(paste0('<iframe src="', file.path(params$result_folder, "heatmap_top1k.pdf"), '" width="100%" height="600px"></iframe>'))
```

# Differentially methylated probes and regions

## Differentially methylated probes (DMPs)

```{r DE-test, results='asis'}

if(!params$skip_DM) {
  rownames(coldata) <- coldata$IDAT

  pval = 0.05
  eff_thres = 0.2
  seg_fdr = 0.05
  
  if(!exists("smry")) { 
    smry = list()
    rerun_DML = TRUE
  }else{
    rerun_DML = FALSE
  }
  test_result = list()
  test_result_sig = list()
  test_result_dt = list()
  merged = list()
  merged_sig = list()
  merged_dt = list()
  merged_plt = list()
  warning_flag = c()
  
  for(comp in params$contrast_pairs){
    if(nrow(coldata[coldata$Group %in% comp[1],])==1 && nrow(coldata[coldata$Group %in% comp[2],])==1) {
      replicates = FALSE
      warning_flag <- c(warning_flag, paste(comp, collapse = " vs "))
    }else{
      replicates = TRUE
    }
    
    betas.sub = betas[,coldata$Group %in% comp]
    coldata.sub = coldata[coldata$Group %in% comp,]
    
    coldata.sub$Group <- factor(coldata.sub$Group, levels = comp)
    
    betas_ok = (checkLevels(betas.sub, coldata.sub$Group))
    comp_name = paste(comp, collapse = " vs ")
    cat("<h4>", comp_name, "</h4>")
    cat("We removed probes missing a non-NA value for any factor level (", paste0(unique(coldata.sub$Group), collapse = " or "), " in this contrast). The remaining ", sum(betas_ok), " out of ", nrow(betas.sub), " probes were used for DM analysis.", sep = "")
  
    betas.sub = betas.sub[betas_ok,]
    if(params$test_mode) betas.sub = betas.sub[1:2000,] # for testing
    
    if(is.null(params$batch_correction_method)) {
      if(rerun_DML) smry[[comp_name]] = DML(betas.sub, fm = ~Group, meta = coldata.sub)
      test_result[[comp_name]] = summaryExtractTest(smry[[comp_name]])
      
      if(replicates) {
        test_result_sig[[comp_name]] = test_result[[comp_name]] %>% filter(.[[5]] < pval, .[[7]] > eff_thres) %>% arrange(desc(.[[7]]))
      }else{
        test_result_sig[[comp_name]] = test_result[[comp_name]] %>% filter(.[[7]] > eff_thres) %>% arrange(desc(.[[7]]))
      }
    }else{
      if(rerun_DML) smry[[comp_name]] = DML(betas.sub, fm = ~Group+Batch, meta = coldata.sub)
      test_result[[comp_name]] = summaryExtractTest(smry[[comp_name]])
      
      if(replicates) {
        pval_index <- grep("^Pval_Group",colnames(test_result[[comp_name]]))
        eff_index <- grep("Eff_Group",colnames(test_result[[comp_name]]))
        test_result_sig[[comp_name]] = test_result[[comp_name]] %>% filter(.[[pval_index]] < pval, .[[eff_index]] > eff_thres) %>% arrange(desc(.[[pval_index]]))
      }else{
        test_result_sig[[comp_name]] = test_result[[comp_name]] %>% filter(.[[eff_index]] > eff_thres) %>% arrange(desc(.[[eff_index]]))
      }
    }
    
    test_result_dt[[comp_name]] = DT::datatable(data = test_result_sig[[comp_name]][1:100,], caption = paste(comp_name, "Top 100 Differential Methylated CpGs"), rownames = F)
  
    if(replicates) {
      cat("There were ", nrow(test_result_sig[[comp_name]]), " differentially methylated CpGs with a p-value < ", pval, " and an effective size > ", eff_thres, ".<br>", sep = "")
    }else{
      cat("***Note that there was no replicate for either ", comp[1], " or ", comp[2], ". So P-value was not calculated for this comparison. All the ", nrow(test_result_sig[[comp_name]]), " CpGs with an effective size > ", eff_thres, " were saved to the results tables.***<br>", sep = "")
    }
    cat("Up-regulated in", comp[2], ":", nrow(test_result_sig[[comp_name]] %>% filter(.[[3]]>0)), "<br>")
    cat("Down-regulated in", comp[2], ":", nrow(test_result_sig[[comp_name]] %>% filter(.[[3]]<0)), "<br>")
  
    if(!dir.exists(file.path(params$result_folder, "DMP/"))) dir.create(file.path(params$result_folder, "DMP/"), recursive = TRUE)
    write.csv(test_result_sig[[comp_name]], file = paste0(params$result_folder, "/DMP/DMP_", gsub("\\s", "_", comp_name), ".csv"))
    
    merged[[comp_name]] = DMR(betas.sub, smry[[comp_name]], colnames(attr(smry[[comp_name]], "model.matrix"))[2]) # merge CpGs to regions
    if(replicates) {
      merged_sig[[comp_name]] = merged[[comp_name]] %>% dplyr::filter(abs(Seg_Est) > eff_thres) %>% dplyr::filter(Seg_Pval_adj < seg_fdr) %>% arrange(desc(abs(Seg_Est)))
    }else{
      merged_sig[[comp_name]] = merged[[comp_name]] %>% dplyr::filter(abs(Seg_Est) > eff_thres) %>% arrange(desc(abs(Seg_Est)))
    }
    merged_dt[[comp_name]] = DT::datatable(data = unique(merged_sig[[comp_name]][grep("chr",merged_sig[[comp_name]]$Seg_Chrm),1:7]), caption = paste(comp_name, "Differential Methylated Regions (DMRs)"), rownames = F)
    merged_plt[[comp_name]] = unique(merged_sig[[comp_name]][grep("chr",merged_sig[[comp_name]]$Seg_Chrm),1:7])
  
    if(!dir.exists(file.path(params$result_folder, "DMR/"))) dir.create(file.path(params$result_folder, "DMR/"), recursive = TRUE)
    write.csv(merged_sig[[comp_name]], file = paste0(params$result_folder, "/DMR/DMR_", gsub("\\s", "_", comp_name), ".csv"))
  }
}

```

### Top 100 significant DMPs

The tables below show top 100 significant DMPs. Please find the full significant DMP lists in the `DMP` folder.

**Note:** With the exception of the Intercept, there are four groups of columns, each starting with "Est_", "Pval_", "FPval_", and "Eff_" as prefix. Here are what they represent:

- Est: The slope estimate (aka the β coefficient, not to be confused with the DNA methylation β-value though) for continuous variable. DNA methylation difference of the current level with respect to the reference level for nominal contrast variables. There is a special column named Est_X.Intercept. It corresponds to the base-level methylation of the reference.

- Pval: The unadjusted p-values of t-testing the slope. This represents the statistical significance of the methylation difference. The Pval_X.Intercept tests whether the reference level is significantly different from zero.

- FPval: The unadjusted p-value of the F-test contrasting the full model against a reduced model with the labeled contrast variable held out. Note that "Pval_" and "FPval_" are equivalent when the contrast variable is a 2-level factor, i.e., in the case of a pairwise comparison.

- Eff: The effect size of each normial contrast variable. This is equivalent to the maximum slope subtracted by the minimum level including the reference level (0).

```{r DMP tables}
if(!params$skip_DM) {
  htmltools::tagList(setNames(test_result_dt, NULL))
}
```

### Barplots of significant DMPs

```{r barplot-function}
DM_barplots <- function(plot_data, title){
  space = (max(plot_data$Hypermethylation)-min(plot_data$Hypomethylation))*0.15/2
  bp_ratio = nrow(plot_data)/(max(plot_data$Hypermethylation)+max(plot_data$Hypomethylation))
  ggplot(plot_data, aes(x = Contrast)) +
    geom_bar(aes(y = Hypermethylation, fill = "Hypermethylation"), stat = "identity", position = "identity") +
    geom_bar(aes(y = -Hypomethylation, fill = "Hypomethylation"), stat = "identity", position = "identity") +
    geom_text(aes(label = Hypermethylation, y = Hypermethylation), vjust = -0.5) +
    geom_text(aes(label = Hypomethylation, y = -Hypomethylation), vjust = 1.5) +
    scale_fill_manual(values = c("Hypermethylation" = "#FC8D62", "Hypomethylation" = "#66C2A5"),
                      breaks = c("Hypermethylation", "Hypomethylation"),
                      guide = guide_legend(title = "Methylation Status")) +
    labs(title = title) +
    theme_minimal() +
    theme(panel.grid.major.x = element_blank(),
          panel.border = element_rect(colour = "black", fill = NA),
          legend.text = element_text(size = 10),
          legend.title = element_text(size = 11),
          axis.text.x = element_text(angle = 45, hjust = 1),
          axis.text = element_text(size = 8),
          axis.text.y = element_blank(),   # Remove y-axis labels
          axis.title.y = element_blank()) +
    ylim(-max(plot_data$Hypomethylation)-space*15,max(plot_data$Hypermethylation)+space*15) 
}
```

```{r barplot-DMPs}
if(!params$skip_DM) {
  plot_data <- lapply(test_result_sig, function(x) table(as.vector(x[[3]]>0))) %>% do.call(rbind, .)
  colnames(plot_data) <- c("Hypomethylation","Hypermethylation")
  plot_data <- rownames_to_column(as.data.frame(plot_data), var = "Contrast")
  plot_data$Contrast <- factor(plot_data$Contrast, levels = names(test_result_sig))
  DM_barplots(plot_data, "Number of DMPs")
}
```

<!-- ### Volcano plots of all CpGs -->

<!-- ```{r volcano-function} -->
<!-- plot_volcano_dmp <- function(title, res, p.thres, fc.thres) { -->
<!--   res <- res %>% as.data.frame() -->
<!--   res_up <- res[res[, 3] > fc.thres & res[, 5] < p.thres, ] %>% na.omit() -->
<!--   res_down <- res[res[, 3] < -fc.thres & res[, 5] < p.thres, ] %>% na.omit() -->
<!--   # Prepare the base plot -->
<!--   vplot <- ggplot(res, aes(.data[[names(res)[3]]], -log10(.data[[names(res)[5]]]), text = Probe_ID)) + -->
<!--     geom_point(size = 2) + -->
<!--     geom_hline(yintercept = -log10(p.thres), linetype = "longdash", colour = "grey", size = 1) + -->
<!--     geom_vline(xintercept = fc.thres, linetype = "longdash", colour = "#BE684D", size = 1) + -->
<!--     geom_vline(xintercept = -fc.thres, linetype = "longdash", colour = "#2C467A", size = 1) + -->
<!--     labs(title = title, subtitle = "Volcano plot") + -->
<!--     ylab("-log10 P") + -->
<!--     xlab("DNA Methylation Difference") + -->
<!--     theme_bw() + -->
<!--     scale_x_continuous(limits = c(-0.8, 0.8), oob = scales::squish) -->

<!--   # Highlight points based on thresholds -->
<!--   if (nrow(res_up) > 0) { vplot <- vplot + geom_point(data = res_up, aes(x = .data[[names(res_up)[3]]], y = -log10(.data[[names(res_up)[5]]])), color = 'red') } -->
<!--   if (nrow(res_down) > 0) { vplot <- vplot + geom_point(data = res_down, aes(x = .data[[names(res_down)[3]]], y = -log10(.data[[names(res_down)[5]]])), color = 'blue') } -->

<!--   interactive_plot <- plotly::ggplotly(vplot, tooltip = "text") -->
<!--   return(interactive_plot) -->
<!-- } -->
<!-- plot_volcano_dmr <- function(title, res, p.thres, fc.thres) { -->
<!--   res <- res %>% as.data.frame() -->
<!--   res_up <- res[res[, "Seg_Est"] > fc.thres & res[, "Seg_Pval_adj"] < p.thres, ] -->
<!--   res_down <- res[res[, "Seg_Est"] < -fc.thres & res[, "Seg_Pval_adj"] < p.thres, ] -->
<!--   # Prepare the base plot -->
<!--   vplot <- ggplot(res, aes(Seg_Est, -log10(Seg_Pval_adj), text = Seg_ID)) + -->
<!--     geom_point(size = 2) + -->
<!--     geom_hline(yintercept = -log10(p.thres), linetype = "longdash", colour = "grey", size = 1) + -->
<!--     geom_vline(xintercept = fc.thres, linetype = "longdash", colour = "#BE684D", size = 1) + -->
<!--     geom_vline(xintercept = -fc.thres, linetype = "longdash", colour = "#2C467A", size = 1) + -->
<!--     labs(title = title, subtitle = "Volcano plot") + -->
<!--     ylab("-log10 Pval Adj") + -->
<!--     xlab("DNA Methylation Difference") + -->
<!--     theme_bw() + -->
<!--     scale_x_continuous(limits = c(-max(res[["Seg_Est"]]), max(res[["Seg_Est"]])), oob = scales::squish) -->

<!--   # Highlight points based on thresholds -->
<!--   if (nrow(res_up) > 0) { vplot <- vplot + geom_point(data = res_up, aes(Seg_Est, -log10(Seg_Pval_adj)), color = 'red') } -->
<!--   if (nrow(res_down) > 0) { vplot <- vplot + geom_point(data = res_down, aes(Seg_Est, -log10(Seg_Pval_adj)), color = 'blue') } -->

<!--   interactive_plot <- plotly::ggplotly(vplot, tooltip = "text") -->
<!--   return(interactive_plot) -->
<!-- } -->
<!-- ``` -->

<!-- ```{r volcano} -->
<!-- if(!params$skip_DM) { -->
<!--   vp_list_dmp <- lapply(names(test_result), function(x) plot_volcano_dmp(x, test_result[[x]], pval, eff_thres)) -->
<!--   htmltools::tagList(setNames(vp_list_dmp, NULL)) -->
<!-- } -->
<!-- ``` -->

## Differentially methylated regions (DMRs)

```{r warning flag, results='asis'}
if(!params$skip_DM) {
  if(length(warning_flag)>0) {
    cat("***Note that for comparison(s) ", paste(warning_flag, sep = ", "), ", neither group had a replicate. So there was no p value could be produced for DMRs. The results below show all DMRs mapping to the CpGs with an effective size > ", eff_thres, ".***<br>", sep = "")
  }
}
```

### Tables of significant DMRs

The tables below show unique significant DMRs. Non-chromosome CpGs were not shown. Please find the full significant DMR lists in the `DMR` folder.

```{r DMR tables}
if(!params$skip_DM) {
  htmltools::tagList(setNames(merged_dt, NULL))
}
```

### Barplots of significant DMRs

```{r barplot-DMRs}
if(!params$skip_DM) {
  plot_data <- lapply(merged_plt, function(x) table(factor(as.vector(x$Seg_Est > 0), levels = c(FALSE, TRUE)))) %>% do.call(rbind, .)
  colnames(plot_data) <- c("Hypomethylation","Hypermethylation")
  plot_data <- rownames_to_column(as.data.frame(plot_data), var = "Contrast")
  plot_data$Contrast <- factor(plot_data$Contrast, levels = names(merged_plt))
  DM_barplots(plot_data, "Number of DMRs")
}
```

<!-- ### Volcano plots of all CpGs-merged regions -->

<!-- *Note that the number of significant DMR shown in the volcano plot can be larger than those in the bar-plots and tables above because non-chromosome CpGs are also shown here.* -->

<!-- ```{r volcano regions} -->
<!-- if(!params$skip_DM) { -->
<!--   vp_list_dmr <- lapply(names(merged), function(x) plot_volcano_dmr(x, merged[[x]], seg_fdr, eff_thres)) -->
<!--   htmltools::tagList(setNames(vp_list_dmr, NULL)) -->
<!-- } -->
<!-- ``` -->

# Differentially methylated probe (DMP) enrichment 

## Feature enrichment

Available database sets are listed in the table below. 

```{r database table}
listDBGroups(params$platform) %>% DT::datatable(rownames = FALSE)
```

### Significant DMPs enriched features

The tables below show TFBS enriched with DMPs. Results were filtered with FDR < 0.05 and overlap > 10. 

```{r enrichment test 1}
if(!params$skip_DM) {
  query <- lapply(test_result_sig, function(x) x$Probe_ID)
  cpg_enrichment_result <- lapply(query, function(x) testEnrichment(query = x, databases = "TFBS", platform = params$platform))
  cpg_enrichment_result_dt <- lapply(names(cpg_enrichment_result), function(x) cpg_enrichment_result[[x]] %>% dplyr::filter(overlap>10) %>% dplyr::filter(FDR<0.05) %>% dplyr::select(dbname, estimate, test, FDR) %>% DT::datatable(caption = paste(x, "Feature Enrichments")))
  htmltools::tagList(setNames(cpg_enrichment_result_dt, NULL))
}
```

The dot plots below show the top 20 enriched TFBS Dot color indicates odds ratio (OR) in the Fisher's test. 

```{r enrichment dotplot 1, results = 'hide'}
if(!params$skip_DM) {
  lapply(names(cpg_enrichment_result), function(x) KYCG_plotDot(cpg_enrichment_result[[x]], title = paste(x, "Top 20 Feature Enrichments")))
}
```

The tables below show HM enriched with DMPs. Results were filtered with FDR < 0.05 and overlap > 10. 

```{r enrichment test 2}
if(!params$skip_DM) {
  query <- lapply(test_result_sig, function(x) x$Probe_ID)
  cpg_enrichment_result <- lapply(query, function(x) testEnrichment(query = x, databases = "HM", platform = params$platform))
  cpg_enrichment_result_dt <- lapply(names(cpg_enrichment_result), function(x) cpg_enrichment_result[[x]] %>% dplyr::filter(overlap>10) %>% dplyr::filter(FDR<0.05) %>% dplyr::select(dbname, estimate, test, FDR) %>% DT::datatable(caption = paste(x, "Feature Enrichments")))
  htmltools::tagList(setNames(cpg_enrichment_result_dt, NULL))
}
```

The tables below show Mask enriched with DMPs. Results were filtered with FDR < 0.05 and overlap > 10. 

```{r enrichment test 3}
if(!params$skip_DM) {
  query <- lapply(test_result_sig, function(x) x$Probe_ID)
  cpg_enrichment_result <- lapply(query, function(x) testEnrichment(query = x, databases = "Mask", platform = params$platform))
  cpg_enrichment_result_dt <- lapply(names(cpg_enrichment_result), function(x) cpg_enrichment_result[[x]] %>% dplyr::filter(overlap>10) %>% dplyr::filter(FDR<0.05) %>% dplyr::select(dbname, estimate, test, FDR) %>% DT::datatable(caption = paste(x, "Feature Enrichments")))
  htmltools::tagList(setNames(cpg_enrichment_result_dt, NULL))
}
```

The tables below show ABCompartment enriched with DMPs. Results were filtered with FDR < 0.05 and overlap > 10. 

```{r enrichment test 4}
if(!params$skip_DM) {
  query <- lapply(test_result_sig, function(x) x$Probe_ID)
  cpg_enrichment_result <- lapply(query, function(x) testEnrichment(query = x, databases = "ABCompartment", platform = params$platform))
  cpg_enrichment_result_dt <- lapply(names(cpg_enrichment_result), function(x) cpg_enrichment_result[[x]] %>% dplyr::filter(overlap>10) %>% dplyr::filter(FDR<0.05) %>% dplyr::select(dbname, estimate, test, FDR) %>% DT::datatable(caption = paste(x, "Feature Enrichments")))
  htmltools::tagList(setNames(cpg_enrichment_result_dt, NULL))
}
```

The tables below show ImprintingDMR enriched with DMPs. Results were filtered with FDR < 0.05 and overlap > 10. 

```{r enrichment test 5}
if(!params$skip_DM) {
  query <- lapply(test_result_sig, function(x) x$Probe_ID)
  cpg_enrichment_result <- lapply(query, function(x) testEnrichment(query = x, databases = "ImprintingDMR", platform = params$platform))
  cpg_enrichment_result_dt <- lapply(names(cpg_enrichment_result), function(x) cpg_enrichment_result[[x]] %>% dplyr::filter(overlap>10) %>% dplyr::filter(FDR<0.05) %>% dplyr::select(dbname, estimate, test, FDR) %>% DT::datatable(caption = paste(x, "Feature Enrichments")))
  htmltools::tagList(setNames(cpg_enrichment_result_dt, NULL))
}
```

The tables below show ChromHMM enriched with DMPs. Results were filtered with FDR < 0.05 and overlap > 10. 

```{r enrichment test 6}
if(!params$skip_DM) {
  query <- lapply(test_result_sig, function(x) x$Probe_ID)
  cpg_enrichment_result <- lapply(query, function(x) testEnrichment(query = x, databases = "ChromHMM", platform = params$platform))
  cpg_enrichment_result_dt <- lapply(names(cpg_enrichment_result), function(x) cpg_enrichment_result[[x]] %>% dplyr::filter(overlap>10) %>% dplyr::filter(FDR<0.05) %>% dplyr::select(dbname, estimate, test, FDR) %>% DT::datatable(caption = paste(x, "Feature Enrichments")))
  htmltools::tagList(setNames(cpg_enrichment_result_dt, NULL))
}
```

The tables below show MetagenePC enriched with DMPs. Results were filtered with FDR < 0.05 and overlap > 10. 

```{r enrichment test 7}
if(!params$skip_DM) {
  query <- lapply(test_result_sig, function(x) x$Probe_ID)
  cpg_enrichment_result <- lapply(query, function(x) testEnrichment(query = x, databases = "MetagenePC", platform = params$platform))
  cpg_enrichment_result_dt <- lapply(names(cpg_enrichment_result), function(x) cpg_enrichment_result[[x]] %>% dplyr::filter(overlap>10) %>% dplyr::filter(FDR<0.05) %>% dplyr::select(dbname, estimate, test, FDR) %>% DT::datatable(caption = paste(x, "Feature Enrichments")))
  htmltools::tagList(setNames(cpg_enrichment_result_dt, NULL))
}
```

The tables below show CTCFbind enriched with DMPs. Results were filtered with FDR < 0.05 and overlap > 10. 

```{r enrichment test 8}
if(!params$skip_DM) {
  query <- lapply(test_result_sig, function(x) x$Probe_ID)
  cpg_enrichment_result <- lapply(query, function(x) testEnrichment(query = x, databases = "CTCFbind", platform = params$platform))
  cpg_enrichment_result_dt <- lapply(names(cpg_enrichment_result), function(x) cpg_enrichment_result[[x]] %>% dplyr::filter(overlap>10) %>% dplyr::filter(FDR<0.05) %>% dplyr::select(dbname, estimate, test, FDR) %>% DT::datatable(caption = paste(x, "Feature Enrichments")))
  htmltools::tagList(setNames(cpg_enrichment_result_dt, NULL))
}
```

```{r additional enrichment test for selected database}
# test for a selected database (pending function ...)
# dbs <- getDBs(group_nms = listDBGroups(params$platform)[[1]][1], platform = params$platform)
# cpg_enrichment_result <- lapply(query, function(x) testEnrichment(x, databases = dbs, platform = params$platform))
```

## Gene enrichment

### Significant DMPs enriched genes

Test whether the DMP sets are enriched for any genes. Probe-gene distance for association was set 10000 at most. Results were filtered with FDR < 0.05 and overlap > 10. 

```{r gene enrichment}
if(!params$skip_DM) {
  gene_enrichment_result <- lapply(query, function(x) testEnrichment(x, buildGeneDBs(x, max_distance=10000), platform = params$platform))
  gene_enrichment_result_dt <- lapply(names(gene_enrichment_result), function(x) gene_enrichment_result[[x]] %>% dplyr::filter(overlap>10) %>% dplyr::filter(FDR<0.05) %>% dplyr::select(dbname, estimate, gene_name, FDR, nQ, nD, overlap) %>% DT::datatable(caption = paste(x, "Gene Enrichments")))
  htmltools::tagList(setNames(gene_enrichment_result_dt, NULL))
}
```

### Top 20 enriched genes

The dot plots below show the top 20 enriched genes. Dot color indicates odds ratio (OR) in the Fisher's test. 

```{r gene enrichment plot, results = 'hide'}
if(!params$skip_DM) {
  lapply(names(gene_enrichment_result), function(x) KYCG_plotDot(gene_enrichment_result[[x]], label_by = "gene_name", title = paste(x, "Top 20 Gene Enrichments")))
}
```

## Gene-set over-representatoin analysis

Get DMP-associated genes and test whether these genes are enriched for any GO terms. Probe-gene distance for association was set 10000 at most. Results were filtered with FDR < 0.05. DMP-associated gene-sets with a size < 20 were excluded from the analysis. 

The dot plots below show the top 10 enriched GO terms or KEGG pathways. Please find the full results in tables in the `ORA` folder. 

```{r enrichment-dotplot-function}
gs_dotplot <- function(clusterProfResult, fdr_thres, title) {
  # get the top 10 enriched gene sets in each cluster to plot
  res2plot <- clusterProfResult@compareClusterResult %>% dplyr::filter(p.adjust < fdr_thres) %>% dplyr::group_by(Cluster) %>% dplyr::slice_head(n = 10) %>% dplyr::ungroup()
  if (nrow(res2plot)==0) { return(NULL) }
  # calculate gene ratio
  res2plot$GeneRatio <- sapply(strsplit(res2plot$GeneRatio, "/"), function(x) as.numeric(x[1]) / as.numeric(x[2]))
  
  # set the levels of x and y axis
  res2plot$Cluster <- factor(res2plot$Cluster, levels = names(clusterProfResult@geneClusters))
  res2plot$Description <- factor(res2plot$Description, levels = unique(res2plot$Description))
  
  p <- ggplot(res2plot, aes(x = Cluster, y = Description)) +
    geom_point(aes(size = GeneRatio, colour = p.adjust), show.legend = T) + 
    theme_bw() +
    theme(
      axis.text = element_text(colour = "black"),
      axis.text.x = element_text(size = 3, angle = 90, vjust = 0.5, hjust = 0.5),
      axis.text.y = element_text(size = 8),
      legend.title = element_text(size = 12),
      legend.text = element_text(size = 10),
      plot.margin = unit(c(0.5, 0.5, 0.1, 0.1), "cm")
    ) +
    labs(title = title, x = "", y = "", caption = "") +
    scale_colour_gradient(name = "FDR adjusted\n p-value\n", limits = c(0, fdr_thres)) +
    scale_size_continuous(name = "Gene Ratio")
  print(p)
}
```

```{r gene input}
if(!params$skip_DM) {
  ora_genes <- lapply(query, function(x) lapply(buildGeneDBs(x, max_distance = 10000), function(y) attributes(y)$gene_name) %>% unlist(use.names = FALSE) %>% unique())
}
```

```{r functional enrichment}
if(!params$skip_DM) {

  if(!dir.exists(file.path(params$result_folder, "ORA"))) dir.create(file.path(params$result_folder, "ORA"))

  if (params$species == "human") {
    gsorDb <- "org.Hs.eg.db"
    keggOrg <- 'hsa'
    msigdbr_species <- "Homo sapiens"
  } else if (params$species == "mouse") {
    gsorDb <- "org.Mm.eg.db"
    keggOrg <- 'mmu'
    msigdbr_species <- "Mus musculus"
  } else if (params$species == "rat") {
    gsorDb <- "org.Rn.eg.db"
    keggOrg <- 'rno'
    msigdbr_species <- "Rattus norvegicus"
  } else {
    stop("Species not supported. ")
  }

  deGenes_ora <- ora_genes[sapply(ora_genes, length) >= 20]

  enrichGO_res <- list()
  go_cat <- c('BP','CC','MF')

  if (!all(sapply(deGenes_ora, length) == 0)) {
    # run GO enrichment analysis
    enrichGO_res$all <- lapply(go_cat, function(x) clusterProfiler::compareCluster(geneCluster = deGenes_ora, fun = 'enrichGO', keyType = 'SYMBOL', OrgDb = gsorDb, ont = x, readable = TRUE)) %>% setNames(go_cat)
    # remove null elements
    enrichGO_res$all <- if (all(sapply(enrichGO_res$all, is.null))) NULL else enrichGO_res$all[!sapply(enrichGO_res$all, is.null)]
  }
  if (!is.null(enrichGO_res$all)) {
    # plot the enriched GO terms
    lapply(names(enrichGO_res$all), function(x) gs_dotplot(enrichGO_res$all[[x]], params$ora_fdr_thres, paste0("Top 10 enriched GO ", x, " terms per contrast"))) %>% invisible()
    # save the enriched GO terms
    lapply(names(enrichGO_res$all), function(x) lapply(names(deGenes_ora), function(y) readr::write_csv(subset(enrichGO_res$all[[x]]@compareClusterResult, Cluster==y & p.adjust<params$ora_fdr_thres), file = paste0(file.path(params$result_folder, "ORA"), "/ORA_GO_", x, "_", gsub(" ", "_", y), "_padj", sub("0\\.", "", as.character(params$ora_fdr_thres)), ".csv")))) %>% invisible()
  }

  enrichKEGG_res <- list()

  if (!all(sapply(deGenes_ora, length) == 0)) {
    # map gene names to entrez ids
    deEntrez <- lapply(deGenes_ora, function(x) bitr(x, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = gsorDb)$ENTREZID)
    # run KEGG enrichment analysis
    enrichKEGG_res$all <- clusterProfiler::compareCluster(geneCluster = deEntrez, fun = 'enrichKEGG', organism = keggOrg)
  }
  if (!is.null(enrichKEGG_res$all)) {
    # plot the enriched KEGG pathways
    gs_dotplot(enrichKEGG_res$all, params$ora_fdr_thres, "Top 10 enriched KEGG pathways per contrast")
    # map entrez ids back to gene names
    enrichKEGG_res$all@compareClusterResult$geneID <- lapply(enrichKEGG_res$all@compareClusterResult$geneID, function(x) unlist(strsplit(x, "/")) %>% bitr(fromType = "ENTREZID", toType = "SYMBOL", OrgDb = gsorDb) %>% dplyr::select(SYMBOL) %>% unlist() %>% paste(collapse = "/")) %>% unlist()
    # save the enriched KEGG pathways
    lapply(names(deGenes_ora), function(x) readr::write_csv(subset(enrichKEGG_res$all@compareClusterResult, Cluster==x & p.adjust<params$ora_fdr_thres), file = paste0(file.path(params$result_folder, "ORA"), "/ORA_KEGG_", x, "_padj", sub("0\\.", "", as.character(params$ora_fdr_thres)), ".csv"))) %>% invisible()
  }
}
```

# Metagene methylation levels

```{r metagene}
if(!params$skip_DM) {
  betas_meta <- betas
  rownames(betas_meta) <- gsub("_.*", "", rownames(betas_meta))
  KYCG_plotMeta(betas_meta)
}
```

# Track view with heatmap

We are able to visualize the methylation level across a given genomic region, around a given gene, or around a given probe. Below shows an example of the visualization. Please contact us further if you would like to visualize certain genomic regions. 

The vertical red line on chromosome cytoband indicates the location of the target gene/region. The black boxes are gene bodies. The red boxes inside are exons. There could be multiple gene bodies shown if they are overlapped within the same genomic region. The plot usually shows 2kb flanking regions of the target gene/region. 

```{r track view with heatmap}
visualizeGene('DNMT1', betas)
# visualizeRegion('chr15', 96343814, 96544301, betas, platform="EPIC", show.probeNames = FALSE)
# visualizeGene('DNMT1', betas, platform="EPIC", genome = "hg38")
# visualizeProbes(c("cg00000165"), betas, platform="EPIC", genome = "hg38", upstream = 2000, dwstream = 2000)
```

------------------------------------------------------------------------

# Copy Number Variation

```{r cnv, results='asis'}
for (s in names(segs)) {
  cat("\n\n", s, "(", as.character(coldata[coldata$IDAT==s,"Group"]), ")\n\n")
  # visualizeSegments(segs[[s]]) %>% print()
  visualizeSegments(segs[[s]], to.plot=paste0("chr", 1:22)) %>% print()
}
```

```{r save DML results}
if(!params$skip_DM && !params$test_mode) save(list = ls()[-grep("params", ls())], file = paste0(params$result_folder, ".RData"))
```

# Session info

The output from running 'sessionInfo' is shown below and details all packages and version necessary to reproduce the results in this report.

```{r session info}
sessionInfo()
```

